\scp{\tsc{Timor-Babar}}{B-02}

\noindent{\wg\st{2pt}\begin{xltabular}{\linewidth}
{lllll>{\raggedright\arraybackslash}X}
\lsptoprule
%Language	&	ISO	&	Glotto.	&	banana	&	Etymon	&		\\	\midrule
%\endfirsthead
Language	&	ISO	&	Glotto.	&	banana	&	Etymon	&	Source	\\	\midrule\hhline{~}
\endhead
\ili{Helong}	&	heg	&	helo1243	&	bua	&	\it{misc.}	&	\cite{BalleCameron2014} (PMP *buaq `fruit')	\\
\ili{Tii}	&	txq	&	tiii1241	&	hundi	&	*punti	&	\cite[197]{Jonker1908}	\\
\ili{Lole}	&	llg	&	lole1239	&	hundi	&	*punti	&	\cite{Zacharias2014}	\\
\ili{Termanu}	&	twu	&	term1237	&	huni	&	*punti	&	\cite[197]{Jonker1908}	\\
\ili{Bilbaa}	&	bpz	&	bilb1242	&	huni	&	*punti	&	\cite[197]{Jonker1908}	\\
\ili{Landu}	&	(rgu)	&	land1257	&	hundi	&	*punti	&	Edwards fieldnotes	\\
\ili{Rikou}	&	(rgu)	&	ring1244	&	hundi	&	*punti	&	\cite[197]{Jonker1908}	\\
\ili{Dela}-\ili{Oenale}	&	row	&	dela1251	&	hundi	&	*punti	&	\cite[197]{Jonker1908}	\\
\ili{Dengga}	&	dnk	&	deng1253	&	hundi	&	*punti	&	\cite[197]{Jonker1908}	\\
Roi{\Q}is \ili{Amarasi}	&	(aaz)	&	rois1241	&	uri	&	*punti	&	Edwards fieldnotes	\\
\ili{Kotos} \ili{Amarasi}	&	(aaz)	&	koto1251	&	uki	&	*punti	&	Edwards fieldnotes	\\
\ili{Kusa-Manea}	&	(aoz)	&	kusa1252	&	uki	&	*punti	&	Edwards fieldnotes	\\
\ili{Amanuban}	&	(aoz)	&	aman1264	&	uki	&	*punti	&	Edwards fieldnotes	\\
\ili{Molo}	&	(aoz)	&	moll1242	&	uki	&	*punti	&	Edwards fieldnotes	\\
Fatule{\Q}u	&	(aoz)	&			&	uki	&	*punti	&	Edwards fieldnotes	\\
\ili{Baikeno}	&	bkx	&	baik1238	&	uki	&	*punti	&	Edwards fieldnotes	\\
Amfo{\Q}an	&	(aoz)	&	amfo1237	&	uki	&	*punti	&	Edwards fieldnotes	\\
E. \ili{Tetun}	&	(tet)	&	east2473	&	hudi	&	*punti	&	\cite[88]{Morris1984}	\\
\ili{Fehan} \ili{Tetun}	&	(tet)	&	sout2898	&	hudi	&	*punti	&	\cite{vanKlinken1995}	\\
\ili{Habun}	&	hbu	&	habu1241	&	muku	&	{\#}muku	&	\cite{Dawson2014}	\\
\ili{Kairui-Midiki}	&	krd	&	kair1265	&	hudi	&	*punti	&	\cite{Dawson2014}	\\
Waima{\Q}a	&	wmh	&	waim1252	&	hudi	&	*punti	&	\cite{HimmelmannFreitas2006}	\\
\ili{Naueti}	&	nxa	&	naue1237	&	hudi 	&	*punti	&	\cite{Veloso2016}	\\
\ili{Idate}	&	idt	&	idat1237	&	muu	&	{\#}muku	&	\cite{Dawson2014}	\\
\ili{Galolen}	&	gal	&	galo1243	&	muʔu	&	{\#}muku	&	\cite{CristoReiGrimes2011}	\\
\ili{Galolen}, \ili{Talur}	&	ilw	&	talu1239	&	muʔu	&	{\#}muku	&	\cite[120]{Hinton2000}	\\
\ili{Tugun}	&	tzn	&	tugu1245	&	muʔu	&	{\#}muku	&	\cite[120]{Hinton2000}	\\
Ili{\Q}uun	&	ilu	&	iliu1237	&	muu	&	{\#}muku	&	\cite[128]{DeJosselinDeJong1947}	\\
\ili{Perai}	&	wet	&	pera1257	&	muu	&	{\#}muku	&	\cite[120]{Hinton2000}	\\
\ili{Aputai}	&	apx	&	aput1237	&	muu	&	{\#}muku	&	\cite[120]{Hinton2000}	\\
Dadu{\Q}a	&	(adb)	&	dadu1237	&	muu	&	{\#}muku	&	\cite{Penn2006}	\\
\ili{Rasua}	&	(adb)	&			&	muʔu	&	{\#}muku	&	Grimes fieldnotes	\\
\ili{Raklungu}	&	(adb)	&			&	muʔu	&	{\#}muku	&	Grimes fieldnotes	\\
\ili{Hresuk}	&	(adb)	&			&	muʔu	&	{\#}muku	&	\cite{Boarccaech2013}	\\
\ili{Makuva}	&	lva	&	maku1277	&	utu-ke	&	*punti	&	\cite{vanEngelenhoven2009}	\\	\hhline{~}
(\ili{Lovaia})	&		&		&		&		&	\hp{\,}	\\
E. \ili{Damer}, 	&	(dmr)	&	(east2472)	&	bi	&	*biak	&	Edwards fieldnotes	\\	\hhline{~}
\ili{Ilih}&&&&& \hp{\,}\\
E. \ili{Damer}, 	&	(dmr)	&	(east2472)	&	vu	&	\it{misc.}	&	\cite[412]{Taber1993}	\\	\hhline{~}
\ili{Wulur}&&&&& \hp{\,}\\
\ili{Roma}	&	rmm	&	roma1332	&	wui	&	\it{misc.}	&	\cite[412]{Taber1993}	\\
\ili{Kisar}	&	kje	&	kisa1266	&	muʔu	&	{\#}muku	&	\cite[412]{Taber1993}	\\
\ili{Wetan}	&	(lex)	&	weta1243	&	udi	&	*punti	&	\cite{DeJosselinDeJong1987}	\\
\ili{Luang}	&	(lex)	&	luan1263	&	udi	&	*punti	&	\cite[412]{Taber1993}	\\
\ili{Serua}	&	srw	&	seru1245	&	wia	&	*biak	&	\cite[412]{Taber1993}	\\
\ili{Teun}	&	tve	&	teun1241	&	fiwa	&	*biak	&	\cite[412]{Taber1993}	\\
\ili{Nila}	&	nil	&	nila1244	&	hia	&	*biak	&	\cite[412]{Taber1993}	\\
W. \ili{Damer}	&	drn	&	west2548	&	uss-o\footnotemark	&	*punti  	&	\cite{ChlenovChlenova2008b}	\\
Dawera-	&	ddw	&	dawe1237	&	(h)urs-ol	&	*punti	&	\cite[171]{Chlenova2002}	\\	\hhline{~}
Daweloor&&&&& \hp{\,}\\
\ili{Dai}	&	dij	&	daii1240	&	diamɲon	&	\it{misc.}	&	\cite[412]{Taber1993}	\\
N. \ili{Babar}	&	bcd	&	nort2860	&	udwiya	&	*punti +	&	\cite[412]{Taber1993}	\\	\hhline{~}
	&		&		&		&	*biak	&	\hp{\,}	\\
C. \ili{Marsela}	&	mxz	&	cent2072	&	ut	&	*punti	&	\cite[412]{Taber1993}	\\
SE. \ili{Babar}	&	vbb	&	sout2883	&	uty	&	*punti	&	\cite[412]{Taber1993}	\\
\ili{Serili}	&	sve	&	seri1255	&	ut	&	*punti	&	\cite[412]{Taber1993}	\\
E. \ili{Marsela}	&	vme	&	east2462	&	ut	&	*punti	&	\cite[412]{Taber1993}	\\
\ili{Emplawas}	&	emw	&	empl1237	&	uty-o	&	*punti	&	Edwards fieldnotes	\\
\ili{Imroing}	&	imr	&	imro1237	&	uti-e	&	*punti	&	\cite[412]{Taber1993}	\\
\ili{Tela-Masbuar}	&	tvm	&	tela1241	&	uti-e	&	*punti	&	\cite[412]{Taber1993}	\\
\ili{Selaru}	&	slu	&	sela1259	&	kwe	&	\it{misc.}	&	DD	\\
\ili{Seluwasan}	&	sws	&	selu1242	&	ʤabe	&	\it{misc.}	&	\cite{Conner2012}	\\
\ili{Makatian}	&	{-}{-}{-}	&	maka1307	&	ʤave	&	\it{misc.}	&	\cite{Conner2012}	\\
\lspbottomrule
\end{xltabular}}
\footnotetext{We thank a reviewer for pointing out that West \ili{Damer} \it{uss-o}
		and \ili{Dawera-Daweloor} \it{(h)urs-ol} are reflexes of *punti,
		with a reflex of \tsc{Peripheral Babar} **d (from PMP *s/*nt)
		conditioned by the following vowel,
		which had undergone fortition to \it{s}.
		That is: *punti > Proto-\ili{Babar} **uʈi > **udi > Proto-Peripheral \ili{Babar} **uds.
		The same changes are seen in *isi ‘flesh’ > **iʈi > **idi > **ids
		> West \ili{Damer} \it{iss-o}
		and \ili{Dawera-Daweloor} \it{irs-el-ol}, ‘meat,
		flesh’.}