\sscp{Other Papuan}{B-13-03}

\noindent{\wg\st{2pt}\begin{xltabular}{\linewidth}
{llllll>{\raggedright\arraybackslash}X}
\lsptoprule
%Language	&	ISO	&	Glotto.	&	Family	&	banana	&	Etymon	&	Source	\\	\midrule
%\endfirsthead
Language	&	ISO	&	Glotto.	&	Family	&	banana	&	Etymon	&	Source	\\	\midrule\hhline{~}
\endhead
\ili{Moi}	&	mxn	&	moii1235	&	\tsc{W. Bird's Head}	&	o	&	\it{misc.}	&	DD	\\
\ili{Konda}	&	knd	&	kond1303	&	\tsc{\ili{Konda}-Yahadian}				&	mofo	&	\it{misc.}	&	\cite{BerryBerry1987}	\\
\ili{Yahadian}	&	ner	&	yaha1248	&	\tsc{\ili{Konda}-Yahadian}			&	romino,		&	\it{misc.}	&	\cite{BerryBerry1987}	\\
\ili{Inanwatan}	&	szp	&	suab1238	&	\tsc{S. Bird's Head}	&	ˈɸugido	&	*punti	&	\cite{deVries2004}	\\
\ili{Puragi}	&	pru	&	pura1253	&	\tsc{S. Bird's Head}	&	amimi	&	\it{misc.}	&	\cite{deVries2004}	\\
\ili{Kais}	&	kzm	&	kais1235	&	\tsc{S. Bird's Head}	&	rauminu	&	\it{misc.}	&	\cite{BerryBerry1987}	\\
\ili{Kokoda}	&	xod	&	koko1265	&	\tsc{S. Bird's Head}	&	udi(a)	&	*punti	&	\cite{deVries2004}	\\
\ili{Kaburi}	&	uka	&	kabu1249	&	\tsc{S. Bird's Head}	&	hamimo	&	\it{misc.}	&	\cite{BerryBerry1987}	\\
\ili{Kemberano}	&	bzp	&	kemb1235	&	\tsc{S. Bird's Head}	&	undi	&	*punti	&	\cite{Voorhoeve1984}	\\
\ili{Maybrat}	&	ayz	&	maib1239	&	\tsc{\ili{Maybrat}-Karon}	&	apit	&	\it{misc.}	&	\cite{Dol2007}	\\
\ili{Mor}	&	moq	&	morb1239	&	Isolate	&	hut(a)	&	*punti	&	DD	\\
\ili{Moskona}	&	mtj	&	mosk1236	&	\tsc{E. Bird's Head}	&	meni	&	\it{misc.}	&	\cite{Gravelle2010b}	\\
\ili{Meyah}	&	mej	&	meya1236	&	\tsc{E. Bird's Head}	&	menei	&	\it{misc.}	&	\cite{Reesink2002d}	\\
\ili{Sougb}	&	mnx	&	mani1235	&	\tsc{E. Bird's Head}	&	neiʤ	&	\it{misc.}	&	\cite{Reesink2002d}	\\
\ili{Mpur}	&	akc	&	mpur1239	&	Isolate	&	ɸa	&	\it{misc.}	&	\cite{Reesink2002d}	\\
\ili{Mansim}	&		&	mans1260	&	\tsc{\ili{Hatam}-Mansim}	&	wat	&	\it{misc.}	&	\cite{Reesink2002c}	\\
\ili{Hatam}	&	had	&	hata1243	&	\tsc{\ili{Hatam}-Mansim}	&	wid	&	\it{misc.}	&	\cite{Reesink2002d}	\\
\ili{Mairasi}	&	zrs	&	nucl1594	&	\tsc{Mairasi}	&	weʔi	&	*biak	&	DD	\\
\ili{Tunggare}	&	trt	&	tung1293	&	\tsc{Geelvink Bay}	&	oa	&	\it{other}	&	DD	\\
\ili{Yawa}	&	yva	&	nucl1454	&	\tsc{\ili{Yawa}-Saweru}	&	karey	&	*kalauR	&	\cite{JonesPaai1989}	\\
\ili{Saweru}	&	swr	&	sawe1240	&	\tsc{\ili{Yawa}-Saweru}	&	karei	&	*kalauR	&	DD	\\
\ili{Damal}	&	uhn	&	dama1272	&	Isolate	&	kelo	&	*kalauR	&	DD	\\
\ili{Airoran}	&	air	&	airo1242	&	\tsc{Kwerba}	&	hwaam	&	\it{misc.}	&	DD	\\
\ili{Burmeso}	&	bzu	&	burm1264	&	Isolate	&	nibo	&	\it{misc.}	&	DD	\\
\lspbottomrule
\end{xltabular}}