\sscp{\tsc{East Littoral}}{10-01-02}
\citet[110--114]{Collins1983} provides a discussion of the \tsc{East
Littoral} branch of his \tsc{East Piru Bay},
which he sees as also requiring a distinction between *d/*z and *R/*l/*j.
In this section we re-examine the data
and conclude that, despite the complexities,
we can posit a full merger of *d/*z/*R/*l/*j in this node.

Word initially, PMP *d/*z/*l/*R merge as **l in \tsc{East Littoral}.
Sepa has subsequent **l > \it{r}
before a (historical) high vowel, as described above.
Examples are given in \qf{10-04}--\qf{10-07}.
\vspace{1pt}

\noindent{\wg\st{2pt}\begin{tabularx}
{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}
>{\hspace{-\tabcolsep}\enspace}lllll>
{\raggedright\arraybackslash}X}
\lsptoprule
	%&	PMP	&	Sepa	&	Tamilou	&	Tehoru	&	Teluti	&	gloss	\\	\midrule
%\endfirsthead
	&	PMP	&	Sepa	&	Tamilou	&		Tehoru	&	Teluti	&	gloss	\\	\midrule
%\endhead
\tbx{10-04}	&	*\tbr{d}aRaq	&	\tbb{l}aya	&	\tbb{l}aya	&	\tbb{l}aya	&	\tbb{l}aya	&	blood	\\
	&	*\tbr{d}uha	&	\tbr{r}ua	&	\tbb{l}ua	&	\tbb{l}ua	&	\tbb{l}ua	&	two	\\
	&	*\tbr{d}əpa	&	\tbb{l}ea	&	\tbb{l}ea	&	\tbb{l}efa	&		&	fathom	\\
	&	*\tbr{d}ahun	&	\tbb{l}aun	&	\tbb{l}aun	&		&	\tbb{l}aun	&	leaf	\\	\midrule
\tbx{10-05}	
	&	*\tbr{z}alan	&	\tbb{l}aa tina	&	\tbb{l}aa tina	&		&	\tbb{l}aa tina	&	path, way (irr. *l > {\0})	\\	%\hhline{~}
%	&								&									&									&		&									& (irr. *l > {\0}) \\
	&	*\tbr{z}auq	&	\tbb{l}au	&	\tbb{l}au	&		&	\tbb{l}au-o teli	&	far	\\	\midrule\hhline{~}
\tbx{10-06}	&	*\tbr{l}ima	&	\tbr{r}ima	&	\tbb{l}ima	&		&	\tbb{l}ima	&	five	\\
	&	*\tbr{l}aŋit	&	\tbb{l}aant-o	&	\tbb{l}aant-o	&		&	\tbb{l}aant-o	&	sky	\\
	&	*\tbr{l}ayaR	&	\tbb{l}ial-o	&	\tbb{l}eal-o	&	\cg{(layar-o)}	&	\cg{(layar-o)}	&	sail	\\
	&	*qa\tbr{l}əjaw	&	\tbr{r}ia matan	&	\tbb{l}ia matan	&		&	\tbb{l}ia matan	&	sun	\\
	&	**qa\tbr{l}ipan	&	\tbr{r}i{\tl}\tbr{r}ihan-o	&	\mc{2}{l}{ni{\tl}nia\tbb{l}-o \it{(met.)}}&&	centipede	\\	\midrule
\tbx{10-07}	&	*\tbr{R}usuk	&	\tbr{r}usu	&	\tbb{l}usu	&	\tbb{l}usu-n	&	\tbb{l}usu	&	ribs	\\
	&	*\tbr{R}umbia	&	\tbr{r}ipia	&	\tbb{l}upia	&	\tbb{l}upia	&	\tbb{l}ipia	&	sago	\\
	&	*\tbr{R}amut	&	\tbb{l}amon	&	\tbb{l}amon	&		&	\tbb{l}amun	&	root	\\
	&	*\tbr{R}ibu	&	\tbr{r}ihun	&	\tbb{l}ihun	&		&	\tbr{r}ihun	&	thousand 	\\	\hhline{~}
	&							&							&							&		&							&	(Tel. irr. *R > \it{r})	\\
	&	*\tbr{R}umaq	&	\tbr{r}uma	&	\tbr{y}uma	&	uma	&	uma	&	house (irr. *R, except Sepa)	\\
\lspbottomrule\end{tabularx}}

Word medially
and finally *d/*z/*R/*l/*j merge as **l in some words.
Sepa has **l > \it{r} before a high vowel.
Examples are given in \qf{10-08}--\qf{10-12}.
\mbox{}\\

\noindent{\wg\st{2pt}\begin{tabularx}{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lllll>{\raggedright\arraybackslash}X}
\lsptoprule
	%&	PMP	&	Sepa	&	Tamilou	&	Tehoru	&	Teluti	&	gloss	\\	\midrule
%\endfirsthead
	&	PMP	&	Sepa	&	Tamilou	&		Tehoru	&	Teluti	&	gloss	\\	\midrule
%\endhead
\tbx{10-08}
	&	*ma-u\tbr{d}əhi	&	ana	&	ana &	mu\tbb{l}i	&	ana 	&	Tehoru `back', 	\\ \hhline{~}
	&		&	mu{\tl}mu\tbr{r}i-o	&	mu{\tl}mu\tbb{l}i-o	&		&	mu{\tl}mu\tbb{l}i-o	&	others:	\\ \hhline{~}
	&&&&&& `youngest child' \\
	&	*kə\tbr{d}əŋ	&	o\tbb{l}o	&	o\tbb{l}o	&	o\tbb{l}o	&	o\tbb{l}o	&	stand	\\
	&	*tuhu\tbr{d}	&	tu\tbb{l}e	&	tu\tbb{l}e	&		&	tu\tbb{l}e	&	knee	\\	\midrule
\tbx{10-09}	&	*qa\tbr{z}ay	&	wa\tbb{l}a	&	wa\tbb{l}a	&	wa\tbb{l}a-n	&	wa\tbb{l}a	&	jaw	\\
	&	*haRə\tbr{z}an	&	i\tbb{l}an-o	&	i\tbb{l}an-o	&	i\tbb{l}an-o	&	i\tbb{l}an-o	&	ladder	\\	\midrule
\tbx{10-10}	&	*qu\tbr{l}əj	&	u\tbb{l}e	&	u\tbb{l}e	&	u\tbb{l}e	&		&	maggot	\\
	&	**batə\tbr{l}u	&	huto\tbr{r}u	&		&		&	huto\tbb{l}u	&	thick	\\
	&	**ma\tbr{l}ip	&	ma\tbr{r}ih-e	&	ma\tbb{l}i-e	&		&	ma\tbb{l}if-e	&	laugh	\\
	&	*ta\tbr{l}iŋa	&	ti\tbr{r}ina	&	tinna	&		&	ti\tbb{l}ina	&	ear	\\
	&	*bə\tbr{l}i	&	palahe\tbr{r}i	&	palahe\tbb{l}i	&		&	palahe\tbb{l}i-e	&	sell	\\
	&	*ma-kə\tbr{l}əm	&	kamo\tbb{l}on-o	&	kamo\tbb{l}on-o	&		&	kamo\tbb{l}on-o	&	dark	\\
	&	*ka\tbr{l}aw	&	manʤa\tbb{l}a	&	manya\tbb{l}a	&	manʤa\tbb{l}a	&		&	hornbill	\\
	&	*da\tbr{l}əm	&		&	la\tbb{l}in	&		&	la\tbr{r}in	&	inside (Tel. irr. *l > \it{r})	\\
	&	*ba\tbr{l}iw	&	am-ha\tbb{l}e	&	ha\tbb{l}e	&	ha\tbb{l}e	&		&	opposite side	\\
	&	*sa\tbr{l}aq	&	sa\tbb{l}a	&	sa\tbb{l}a	&	sa\tbb{l}a	&		&	wrong	\\	\midrule
\tbx{10-11}	&	*hadi\tbr{R}i	&	hi\tbr{r}i	&	hi\tbb{l}i	&	hi\tbb{l}i	&	hi\tbb{l}i	&	house post	\\
	&	*pa\tbr{R}ih	&	ha\tbr{r}i	&	ha\tbb{l}i	&		&	fa\tbb{l}i	&	ray	\\
	&	*bu\tbr{R}ah	&	-hu\tbb{l}a	&	-hu\tbb{l}a	&	-hu\tbb{l}a	&		&	spray	\\
	&	*wa\tbr{R}əj	&	wa\tbr{r}it-o	&	wa\tbb{l}it-o	&	wa\tbb{l}it-o	&	wait-o	&	vine, rope	\\
	&	*wahiyə\tbr{R}	&	wae\tbb{l}-o	&	wae\tbb{l}-o	&		&	wae\tbb{l}-o	&	water	\\
	&	*laya\tbr{R}	&	lia\tbb{l}-o	&	lea\tbb{l}-o	&	(layar-o)	&	(layar-o)	&	sail	\\
	&	**niwə\tbr{R}	&	nuwe\tbb{l}-o	&	nuwe\tbb{l}-o	&	nuwe\tbr{r}-o	&	nuwe\tbb{l}-o	&	coconut (Teh. irr. *R>\it{r})	\\	\midrule
\tbx{10-12}	&	*pi\tbr{j}a	&	hi\tbb{l}a	&	hi\tbb{l}a	&	fi\tbb{l}a	&	fi\tbb{l}a	&	how much?	\\
	&	*hua\tbr{j}i	&	wa\tbr{r}i	&	wa\tbb{l}i-n	&	wa\tbb{l}i-n	&	wa\tbb{l}i	&	ySib (same sex)	\\
	&	*pa\tbr{j}ay	&	ha\tbb{l}a	&	ha\tbb{l}a	&	fa\tbb{l}a	&	fa\tbb{l}a	&	rice	\\
	&	*su\tbr{j}a	&	su\tbb{l}a	&	su\tbb{l}a	&	su\tbb{l}a	&		&	stake trap	\\
\lspbottomrule\end{tabularx}\mbox{}\\}

In another set of words,
word medial or final *R/*l/*j > **y in Tamilou, Tehoru, and Teluti.
In some cases **y > {\0},
after triggering adjacent *u > \it{i} (alternately conceptualised
as **yu/**uy > \it{i}).
While in other cases,
**y is retained unchanged or becomes \it{ʤ} through glide fortition.
Sepa also shows *R/*l/*j > \it{y} in some of these words,
though in others it has *R/*l/*j > **l (with subsequent **l >
\it{r} before high vowels.
Examples are given in \qf{10-13}--\qf{10-15}.
\mbox{}\\

\noindent{\wg\st{6pt}\begin{tabularx}{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lllll>{\raggedright\arraybackslash}X}
\lsptoprule
	%&	PMP	&	Sepa	&	Tamilou	&	Tehoru	&	Teluti	&	gloss	\\	\midrule
%\endfirsthead
	&	PMP	&	Sepa	&	Tamilou	&		Tehoru	&	Teluti	&	gloss	\\	\midrule
%\endhead
\tbx{10-13}	&	*qu\tbr{l}u	&	u\tbr{r}u	&	u\tbr{y}u	&	u\tbr{ʤ}u	&	u\tbr{y}u	&	head	\\
	&	*bu\tbr{l}u	&	hu\tbr{r}u-n	&	hu\tbr{y}u-n	&	hu\tbr{ʤ}u-m	&	hu\tbr{i}-n	&	body hair	\\
	&	*ti\tbr{l}u/*tu\tbr{l}i	&	tui-o	&	tui-o	&	ti\tbr{l}u	&		&	earwax	\\
	&	*bu\tbr{l}an	&	h\tbr{i}an-o	&	h\tbr{i}an-o	&	h\tbr{i}an-o	&	h\tbr{i}an-o	&	moon	\\
	&	*tə\tbr{l}u	&	to\tbr{r}u	&	tou	&	to\tbr{i}	&	to\tbr{i}	&	three	\\
	&	*qatə\tbr{l}uR	&	te\tbr{r}un	&	te\tbr{l}un	&	te\tbr{i}-n	&	te\tbr{i}-n	&	egg	\\
	&	*wa\tbr{l}u	&	wa\tbr{r}u	&	wa\tbr{ʤ}o	&		&	wa\tbr{ʤ}u	&	eight	\\
	&	*za\tbr{l}an	&	laa tina	&	laa tina	&		&	laa tina	&	path, way	\\
	&	**ma\tbr{l}abaw	&	m\tbr{l}aha	&	ma\tbr{y}(ah)a	&	ma\tbr{y}aha	&	ma\tbr{y}aha	&	mouse, rat	\\
	&	*kumba\tbr{l}	&	upa-o	&	upa\tbr{y}-o	&	upa\tbr{y}-o	&		&	palm stem, PMP `dry sago branch'	\\	\midrule
\tbx{10-14}	&	*u\tbr{R}at	&	u\tbr{l}at-o	&	u\tbr{y}at-o	&	u\tbr{y}at-o	&	u\tbr{l}at-o	&	vein	\\
	&	*baqə\tbr{R}u	&	ho\tbr{r}u-o	&	ho\tbr{y}u-o	&	ho\tbr{i}ʔ-o	&	ho\tbr{i}-o	&	new	\\
	&	*da\tbr{R}aq	&	la\tbr{y}a	&	la\tbr{y}a	&	la\tbr{y}a	&	la\tbr{y}a	&	blood	\\
	&	*qaba\tbr{R}a	&	ha\tbr{y}a	&	ha\tbr{y}a	&	ha\tbr{y}a	&	ha\tbr{y}a	&	carry on shoulder	\\
	&	*na\tbr{R}ah/*na\tbr{r}a	&	na\tbr{y}a	&		&	na\tbr{y}a	&		&	a tree, Pterocarpus indicus	\\
	&	*tu\tbr{R}un	&	tu\tbr{r}u	&	ku\tbr{y}u	&	ku\tbr{i}	&	ku\tbr{i}	&	go down, descend	\\
	&	*busu\tbr{R}	&	hus\tbr{i}-o	&	hus\tbr{i}-o	&	hus\tbr{i}-o	&		&	bow	\\
	&	*kapu\tbr{R}	&	(losa)	&	hau\tbr{l}-o	&	yaf\tbr{i}-o	&	yaf\tbr{i}-o	&	calcium, lime	\\	\midrule
\tbx{10-15}	&	*qapə\tbr{j}u	&	hu\tbr{r}u-n	&	ho\tbr{l}u-n	&	foi-n	&	foi-n	&	bile	\\
	&	*ŋa\tbr{j}an	&	nia-m	&	ni\tbr{y}a-m	&	ni\tbr{y}a-m	&	na\tbr{y}a-lo	&	name	\\
\lspbottomrule\end{tabularx}\mbox{}\\}

\citet[112]{Collins1983} identifies a conditioning environment
which differentiates the words in \qf{10-13} from those in \qf{10-08}.
He proposes that *R/*l/*j > **y between two back vowels \it{a,
o,} and \it{u,} including word finally before the noun marker \it{-o}.
The potential significance of Collins’ generalisation is that
there are no examples of *d/*z > **y.
He thus suggests that *R/*l/*j first merged to **l,
with subsequent **l > **y between back vowels,
after which *d/*z > **l.
If this is correct,
\tsc{East Littoral} would be another branch in which *d/*z does
not merge with *R/*l/*j.

However, \citet[113]{Collins1983} notes that there are a large
number of exceptions in which *R/*l/*j > \it{l} between non-front
vowels, and states that “It may be premature to attempt to draw out convincing
generalisations.” By our count there are eight exceptions in
which *R/*l/*j does not shift to **y,\footnote{
		Exceptions in
		which *R/*l/*j > \it{l} would include: **batəlu ‘thick’,
		*ma-kələm ‘dark’, *kalaw ‘hornbill’,
		*salaq ‘offence, wrongdoing’ (all in [\ref{ex:10-10}]),
		*buRah ‘spray’ \qf{10-11}, *pajay ‘rice’,
		and *suja ‘stake trap’ (both in [\ref{ex:10-12}]).}
and we further note that there are only two potential cases in
which *d/*z occur between non-front vowels -- the environment
in which **l > \it{y} could occur.
(These two words are *kədəŋ > \it{olo} ‘stand’,
and *qazay > \it{wala} ‘jaw’).
We thus tentatively count \tsc{East Littoral} as a branch of
\tsc{Ambon-Seram} in which *d/*z/*R/*l/*j merge to **l,
with a subsequent unconditioned split of **l > **l, **y.

