\sscp{East Damer}{04-06-03}
The East \ili{Damer} language is spoken in six villages on the eastern
part of \ili{Damer} Island.\footnote{
		\ili{Damer} Island is also referred to as ``\ili{Damar} Island''.
		Final \it{e} is preferred in the region.}
Data comes from \citet{Taber1993},
a 193-item word list from \ili{Wulur} village on the southern part of
the Solat/Kolowati bay and a 259-item word list collected by Edwards,
which represents the variety spoken in the village of \ili{Ilih} on the north\ili{east} coast.
Although the level of cognacy is high between these two lists -- about
90{\%} -- they have undergone different sound changes.

In most cases East \ili{Damer} has a merger of *d/*z/*j/*R > \it{r}.
Examples of *d, *R,
*j, *z are given in \qf{04-13}--\qf{04-16}.
*R/*j > \it{r} and *d > \it{r} also occur in \tsc{\ili{Kisar}-Luang}
and \tsc{\ili{Teun}-\ili{Nila}-Serua}, though with slight complications and/or exceptions.

{\gw\st{4pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lllll}
\lsptoprule\hhline{~}
	&	PMP	&	\ili{Ilih}	&	\ili{Wulur}	&	local gl.	&	\\ \midrule
\endhead
\tbx{04-13}	&	*\tbr{d}uha	&	bo-\tbr{r}u	&	wo-\tbr{r}u	&	two	&	\\
	&	*\tbr{d}aləm	&	\tbr{r}alma-n	&	\tbr{r}alma-no	&	inside	&	\\
	&	*\tbr{d}aRaq	&	\tbr{r}aro	&	\tbr{r}ar	&	blood	&	\\
	&	*\tbr{d}uRi	&	\tbr{r}uri(-n)	&	\tbr{r}uri-n(o)	&	bone, thorn	&	\\
	&	*\tbr{d}uyuŋ	&	\tbr{r}uin	&		&	dugong	&	\\
	&	*\tbr{d}iRi	&		&	na-m\tbr{r}iri	&	stand	&	\\
	&	*ha\tbr{d}iRi	&	\tbr{r}il	&		&	pole, post	&	\\
	&	*ma-\tbr{d}iŋ\tbr{d}iŋ	&	\tcg{(nasitu)}	&	ma\tbr{r}i\tbr{r}in	&	cold	&	\\
	&	*bata\tbr{d}	&	bet\tbr{r}a	&		&	corn	&	\\
	&	*lahu\tbr{d}	&	lo\tbr{r}(u)	&	lu\tbr{r}	&	sea	&	\\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}

{\gw\st{4pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lllll}
\lsptoprule\hhline{~}
	&	PMP	&	\ili{Ilih}	&	\ili{Wulur}	&	local gl.	&	\\ \midrule
\endhead
\tbx{04-14}	&	*\tbr{R}umaq	&	\tbr{r}um(o)	&	\tcg{(hawial)}	&	house	&	\\
	&	*\tbr{R}usuk	&	\tbr{r}usuaksi	&		&	ribs	&	\\
	&	*ha\tbr{R}əmaŋ	&	\tbr{r}eman	&		&	sea eel	&	\\
	&	*ha\tbr{R}əzan	&	\tbr{r}e\tbr{r}an	&		&	ladder	&	\\
	&	*ku\tbr{R}ita	&	\tbr{r}it	&		&	octopus	&	\\
	&	*baqə\tbr{R}u	&	na-be\tbr{r}{\tl}bi\tbr{r}u	&	\tbr{r}a-we\tbr{r}{\tl}we\tbr{r}u	&	new	&	\\
	&	*bə\tbr{R}əqat	&	be\tbr{r}tan	&		&	heavy	&	\\
	&	*bu\tbr{R}iq	&		&	n-wu\tbr{r}i	&	bathe	&	\\
	&	*da\tbr{R}aq	&	ra\tbr{r}o	&	ra\tbr{r}	&	blood	&	\\
	&	*du\tbr{R}i	&	ru\tbr{r}i(-n)	&	ru\tbr{r}i-n(o)	&	bone, thorn	&	\\
	&	*di\tbr{R}i	&		&	na-mri\tbr{r}i	&	stand	&	\\
	&	*haba\tbr{R}at	&	ba\tbr{r}to	&		&	west	&	\\
	&	*ña\tbr{R}a	&	na\tbr{r}a-no	&		&	\mc{2}{l}{opposite sex sibling}	\\
	&	*wa\tbr{R}əj	&	a\tbr{r}as	&	a\tbr{r}a	&	rope	&	\\
	&	*ma-i\tbr{R}aq	&	na-me\tbr{r}{\tl}me\tbr{r}	&	me\tbr{r}{\tl}me\tbr{r}	&	red	&	\\
	&	*iku\tbr{R}	&	i\tbr{r}o-n	&	i\tbr{r}uan	&	tail	&	\\
	&	*kulu\tbr{R}	&	ulu\tbr{r}	&		&	breadfruit	&	\\
	&	*laya\tbr{R}	&		&	la\tbr{r}wo	&	sail	&	\\
	&	*liqə\tbr{R}	&	li\tbr{r}a-n	&	li\tbr{r}a-no	&	neck	&	\\
	&	*niu\tbr{R}	&	nu\tbr{r}	&	nu\tbr{r}u	&	coconut	&	\\
	&	*wahiyə\tbr{R}	&	e\tbr{r}u	&	e\tbr{r}	&	water	&	\\
	&	*timu\tbr{R}	&	tim\tbr{r}u	&		&	\ili{east}	&	\\
	&	*qapu\tbr{R}	&	au\tbr{r}	&		&	calcium, lime	&	\\
	&	*qatəlu\tbr{R}	&	telu\tbr{r}	&	te\tbr{r}uan	&	egg	&	\\	\midrule\hhline{~}
\tbx{04-15}	&	*qalə\tbr{j}aw	&	le\tbr{r}o	&	le\tbr{r}o	&	sun	&	\\
	&	*ŋa\tbr{j}an	&	inanna\tbr{r}a-mu	&	inanna\tbr{r}a-no	&	name	&	\\
	&	*hua\tbr{j}i	&	a\tbr{r}i-n	&	a\tbr{r}i	&	\mc{2}{l}{(same sex) younger sibling}	\\
	&	*ŋi\tbr{j}uŋ	&	nu\tbr{r}no-ku	&	nu\tbr{r}ma-no	&	nose	&	\\
	&	*pi\tbr{j}a	&		&	wo-i\tbr{r}e	&	how much	&	\\
	&	*ma-qapə\tbr{j}u	&	na-mir{\tl}mi\tbr{r}u	&		&	bitter	&	\\
	&	*lalə\tbr{j}	&	lala\tbr{r}	&		&	fly (n.)	&	\\
	&	*pusə\tbr{j}	&	us\tbr{r}a-n	&		&	navel	&	\\
	&	*waRə\tbr{j}	&	ara\tbr{s}	&	ara	&	rope	&	*j > \it{s} \\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}

{\gw\st{4pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lllll}
\lsptoprule\hhline{~}
	&	PMP	&	\ili{Ilih}	&	\ili{Wulur}	&	local gl.	&	\\ \midrule
\endhead
\tbx{04-16}	&	*\tbr{z}ukut	&	\tbr{r}ut	&		&	grass	&	\\
	&	*tu\tbr{z}uq	&	n-tu\tbr{r}u	&		&	point at	&	\\
	&	*haRə\tbr{z}an	&	re\tbr{r}an	&		&	ladder	&	\\
	&	**mo\tbr{z}o	&	na-mor{\tl}mo\tbr{r}	&	mor{\tl}mo\tbr{r}	&	green	&	\\
	&	*\tbr{z}alan	&	\tbr{l}alan, lallo	&	\tbr{l}alno	&	path	&	*z > \it{l}	\\
	&	*\tbr{z}auq	&	\tbr{s}osso	&		&	far	&	*z > \it{s}	\\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}

The \ili{Wulur} variety of East \ili{Damer} has *s > \it{h,
s} -- also in \tsc{\ili{Kisar}-Luang} -- but *s is retained unchanged
in the \ili{Ilih} variety of East \ili{Damer}.
Examples are given in \qf{04-17}.

{\gw\st{6pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}llll}
\lsptoprule
\tbx{04-17}	&	PMP	&	\ili{Ilih}	&	\ili{Wulur}	&	local gl.	\\ \midrule
\endfirsthead
\qf{04-17}	&	PMP	&	\ili{Ilih}	&	\ili{Wulur}	&	local gl.	\\
\endhead
	&	*\tbr{s}iwa	&	bos-\tbr{s}i	&	wo-\tbr{s}i	&	nine	\\
	&	*\tbr{s}iku	&	\tbr{s}iuk	&		&	elbow	\\
	&	*\tbr{s}ai	&		&	\tbr{h}eie	&	who	\\
	&	*\tbr{s}aŋa	&	\tbr{s}ane-n	&		&	branch	\\
	&	*a\tbr{s}u	&	a\tbr{s}u	&	a\tbr{h}u	&	dog	\\
	&	*\tbr{s}u\tbr{s}u	&	\tbr{s}u\tbr{s}u-n	&	\tbr{h}u\tbr{h}u-no	&	br\ili{east}	\\
	&	*bəR\tbr{s}ay	&	be\tbr{s}	&		&	paddle (n.)	\\
	&	*Ru\tbr{s}uk	&	ru\tbr{s}uaksi	&		&	rib	\\
	&	*ma-ə\tbr{s}a	&	m\tbr{s}an	&	me\tbr{h}an	&	one	\\
	&	*lə\tbr{s}uŋ	&		&	le\tbr{h}um	&	mortar	\\
	&	*nu\tbr{s}a	&	nu\tbr{s}	&	nu\tbr{h}	&	island	\\
	&	*ŋi\tbr{s}i	&	ni\tbr{s}a-k	&	ŋi\tbr{h}a-no	&	tooth	\\
	&	*pu\tbr{s}əj	&	u\tbr{s}ran	&		&	navel	\\
	&	*bəRa\tbr{s}	&	abrea\tbr{s}	&	awrera	&	hulled rice	\\
\lspbottomrule	
\end{xltabular}\addtocounter{table}{-1}}

PMP *ŋ > \it{ŋ,
n} {\#}{\gap} and *ŋ > \it{n,
m} /{\gap}{\#} in \ili{Wulur} East \ili{Damer},
though both these changes require confirmation from more data.
In other environments, and in \ili{Ilih} East \ili{Damer}, *ŋ/*n > \it{n}.
\ili{Wulur} East \ili{Damer} appears to be the only \tsc{Southwest Maluku} language
which retains *ŋ = \it{ŋ} unchanged in at l\ili{east} some words.
Examples are given in \qf{04-18}.

{\wg\st{4pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lll>{\raggedright\arraybackslash}X}
\lsptoprule
\tbx{04-18}	&	PMP	&	\ili{Ilih}	&	\ili{Wulur}	&	local gl.	\\ \midrule
\endhead
	&	*\tbr{ŋ}isi	&	\tbr{n}isa-k	&	\tbr{ŋ}iha-no	&	tooth \\
	&	*tali\tbr{ŋ}a	&	t\tbr{n}ina-n	&	\tbr{ŋ}ina-no	&	ear (consonant, metathesis, irr. *l > \it{n}; cf. \ili{Leti} \it{tnina} ear, listen)\\
	&	*tali\tbr{ŋ}a	&		&	ga-tli\tbr{n}	&	hear	\\
	&	*\tbr{ŋ}ajan	&	inan\tbr{n}ara-mu	&	inan\tbr{n}ara-no	&	name	\\
	&	*\tbr{ŋ}iju\tbr{ŋ}	&	\tbr{n}ur\tbr{n}o-ku	&	\tbr{n}ur\tbr{m}a-no	&	nose	\\
	&	*la\tbr{ŋ}it	&	lia\tbr{n}ti	&	lia\tbr{n}ti	&	sky	\\
	&	*na\tbr{ŋ}uy	&		&	n-na\tbr{n}i	&	swim	\\
	&	*ha\tbr{ŋ}in	&	a\tbr{n}in, a\tbr{n}ni	&	a\tbr{n}in	&	wind	\\
	&	*sa\tbr{ŋ}a	&	sa\tbr{n}e-n	&		&	branch	\\
	&	*ma-diŋdi\tbr{ŋ}	&		&	mariri\tbr{n}	&	cold	\\
	&	*haRəma\tbr{ŋ}	&	rema\tbr{n}	&		&	sea eel	\\
	&	*duyu\tbr{ŋ}	&	rui\tbr{n}	&		&	dugong	\\
	&	*ləsu\tbr{ŋ}	&		&	lehu\tbr{m}	&	mortar	\\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}

\ili{Ilih} East \ili{Damer} is an important witness within \tsc{Timor-Babar}
as it retains *b = \it{b} in all word positions.
\ili{Wulur} East \ili{Damer} usually has *b > \it{w.} (A variety of realisations
of /w/ occur in \citet{Taber1993} including [v], [ʋ], and [w]).
\ili{Ilih} East \ili{Damer} *b = \it{b} is unlikely due to fortition of **w
> \it{b,} as in most cases *w > {\0}.
Examples of *b are given in \qf{04-19} and *w in \qf{04-20}.

{\wg\st{2pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}llll>{\raggedright\arraybackslash}X}
\lsptoprule
	&	PMP	&	\ili{Ilih}	&	\ili{Wulur}	&	& local gl.	\\ \midrule
\endhead
\tbx{04-19}	&	*\tbr{b}aqəRu	&	na-ber{\tl}\tbr{b}iru	&	ra-wer{\tl}\tbr{w}eru	&	[raverˈveru]	&	new		\\
	&	*\tbr{b}atu	&	\tbr{b}at	&	\tbr{w}atu	&	[ˈwatu]	&	stone		\\
	&	*\tbr{b}əRas	&	a\tbr{b}reas	&	a\tbr{w}rera	&	[avˈrera]	&	hulled rice		\\
	&	*\tbr{b}uaq	&	\tcg{(isin)}	&	\tbr{b}uan	&	[ˈbuan]	&	fruit		\\
	&	*\tbr{b}uhək	&	\tbr{b}uan	&	\tbr{b}uga-no	&	[buˈgano]	&	head hair	(glide fortition **w > g)\\
	&	*\tbr{b}ulan	&	\tbr{b}ullo	&	\tbr{w}ulno	&	[ˈvulnɔ]	&	moon		\\
	&	*\tbr{b}ulu	&	\tbr{b}ulu-m	&	\tbr{w}ulu-m	&	[ˈvulum]	&	body hair		\\
	&	*\tbr{b}əRəqat	&	\tbr{b}ertan	&		&		&	heavy	\\
	&	*\tbr{b}aq\tbr{b}aq	&	\tbr{b}a\tbr{b}a-k	&	\tbr{w}o\tbr{w}a-ro	&	[ˈwowaro]	&	mouth		\\
	&	*\tbr{b}a\tbr{b}uy	&	\tbr{b}a\tbr{b}	&		&		&	pig		\\
	&	*\tbr{b}əRsay	&	\tbr{b}es	&		&		&	paddle (n.)		\\
	&	*\tbr{b}uqaya	&	\tbr{b}u	&		&		&	crocodile		\\
	&	*\tbr{b}uRiq	&		&	n-\tbr{w}uri	&	[n̩ˈvuri]	&	bathe		\\
	&	*ha\tbr{b}aRat	&	\tbr{b}arto	&		&		&	west		\\
	&	*\tbr{b}ala\tbr{b}a\tbr{w}	&	\tbr{b}la\tbr{b}	&	\tbr{w}au	&	[ˈʋau]	&	mouse, rat		\\
	&	*qa\tbr{b}u > 	&	\hp{ }	& \hp{ }	&	\hp{ }&	\hp{ }	\\ \hhline{~}
	&	**li-a\tbr{b}u	&	llia\tbr{b}u	&	liawu	&	[ˈliavu]	&	ash	\\
	&	*tə\tbr{b}uh	&	te\tbr{b}	&		&		&	sugarcane		\\
	&	**si\tbr{b}i	&	si\tbr{b}	&		&		&	chicken		(cf. \ili{Leti} \it{siβi}, \ili{Teun} \it{sifa})\\
	&	**ta\tbr{b}əq	&	ta\tbr{b}e-n	&	tawe-n	&	[ˈtaven]	&	leaf		(cf. \ili{Leti} \it{taβi,} \ili{Rampi} \it{taweʔ} [\citealp[334]{Zobel2020}])\\ \midrule\hhline{~}
\tbx{04-20}	&	*\tbr{w}ahiyəR	&	eru	&	er	&		&	water		\\
	&	*\tbr{w}akaR	&	aharan	&	aharan	&		&	root		\\
	&	*\tbr{w}alu	&	bu-alu	&	\tbr{gw}alu	&		&	eight		\\
	&	*si\tbr{w}a	&	bos-si	&	wo-si	&		&	nine		\\
	&	*qasa\tbr{w}a	&	sa-no	&		&		&	wife		\\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}

East \ili{Damer} appears to have *-aw/*-ay/*a > **a,
but the data is ambiguous due to *a > {\0}, \it{o} /{\gap\#}.
(Some instances of final \it{o} may be suffixes.) There is only
one unambiguous instance of *-aw > \it{a} in the available data.
Examples of final vowel-glide sequences
and *a in final syllables are given in \qf{04-21}--\qf{04-23}.

{\wg\st{6pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}llll}
\lsptoprule
	&	PMP	&	\ili{Ilih}	&	\ili{Wulur}	& local gl.	\\ \midrule
\endhead
\tbx{04-21}	&	*balab\tbr{aw}	&	blab	&	wau	&	mouse, rat	\\
	&	*lak\tbr{aw}	&		&	n-laʔ\tbr{a}t	&	walk	\\
	&	*qaləj\tbr{aw}	&	ler\tbr{o}	&	ler\tbr{o}	&	sun	\\
	&	*qaləj\tbr{aw}	&	ler	&		&	day	\\
	&	*ma-lin\tbr{aw}	&	na-mlin	&		&	calm, still	\\	\midrule\hhline{~}
\tbx{04-22}	&	*bəRs\tbr{ay}	&	bes	&		&	paddle	\\
	&	*mat\tbr{ay}	&		&	m-mat	&	die	\\
	&	*qən\tbr{ay}	&	en	&	\tcg{(matar)}	&	sand	\\	\midrule\hhline{~}
\tbx{04-23}	&	*bu\tbr{a}q	&	pu	&		&	betel nut	\\
	&	*daR\tbr{a}q	&	rar\tbr{o}	&	rar	&	blood	\\
	&	*zal\tbr{a}n	&	lal\tbr{a}n, lall\tbr{o}	&	laln\tbr{o}	&	patj	\\
	&	*lim\tbr{a}	&	bolim	&	wolim	&	five	\\
	&	*kak\tbr{a}k	&	kak\tbr{a}-no	&	ʔaʔ\tbr{a}-no	&	older sibling	\\
	&	*kuRit\tbr{a}	&	rit	&		&	octopus	\\
	&	*ma-qat\tbr{a}q	&	ma{\tl}mat	&		&	ripe	\\
	&	*mat\tbr{a}	&	mat\tbr{a}-k	&	mat\tbr{a}-no	&	eye	\\
	&	*ñaR\tbr{a}	&	nar\tbr{a}-no	&		&	opposite sex sibling	\\
	&	*nus\tbr{a}	&	nus	&	nuh	&	island	\\
	&	*Rum\tbr{a}q	&	rum	&	\tcg{(hawial)}	&	house	\\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}
