\scp{Helong}{04-01}
\ili{Helong} [heg] is spoken in western Timor,
around the area of \ili{Kupang}
and on the island of \ili{Semau}.
\ili{Helong} has been in extensive contact with the \tsc{Rote-Meto}
languages (\srf{04-02}) and shares much vocabulary in
common as a result of borrowing
at multiple stages of their history.
Three varieties of \ili{Helong} can be identified: \ili{Funai} \ili{Helong} spoken
around the area of \ili{Kupang} \citep{Edwards2018b},
\ili{Bolok} \ili{Helong} spoken around the port village of \ili{Bolok} at the far
\ili{east} of the Timor mainland,
and \ili{Semau} \ili{Helong} spoken on \ili{Semau} Island.
Among other differences, \ili{Bolok} often has /r/
where \ili{Semau} has /d/. One example is \ili{Bolok}
\it{dara {\tl} daar}, \ili{Semau} \it{dada {\tl} daad} ‘sit’.

\ili{Helong} has *p > {\0}, \it{h} /(V\tsc{+fr}
and *p > {\0} elsewhere.
It can thus be placed within \tsc{Timor-Babar}.
Examples in addition to those in \trf{04-01}
are given in \qf{Hel-p}.

{\wg\st{6pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lll>{\raggedright\arraybackslash}X}
\lsptoprule
						&	PMP				&	\ili{Funai}	&	\ili{Semau}	&	local gl.	\\ \midrule
%\endfirsthead
%						&	PMP				&	\ili{Funai}	&	\ili{Semau}	&	local gl.	\\
%\endhead
\tbx{Hel-p}	&	*\tbr{p}ajay		&	ale							&	ale			&	rice plant	\\
%						&	*\tbr{p}aRih		&	\mc{2}{c}{ano}	&fungal infection (unspecified \ili{Helong} from \citet[161]{Jonker1908}	\\
%						&	*\tbr{p}utput		&	\mc{2}{c}{pupu}	&	blowpipe (unspecified variety of \ili{Helong} from \citealt[169]{Jonker1908})	\\
%						&	*\tbr{p}ija			&	ila					&	ila					&	how many?	\\
%						&	*\tbr{p}itu			&	itu					&	itu					&	seven	\\
						&	*\tbr{p}uqun		&	uun					&	uun					&	bamboo (PMP `tree trunk, source')	\\
						&	*\tbr{p}ənuq		&	inu					&	inu					&	full	\\
%						&	*\tbr{p}əRəq		&	\mc{2}{c}{hele}	&	squeeze out (unspecified variety of \ili{Helong} from \citealt[169]{Jonker1908})	\\
						&	*\tbr{p}aRih		&	\mc{2}{c}{ai}		&	stingray (unspecified \ili{Helong} from \citet[153]{Jonker1908}	\\
						&	*\tbr{p}əñu			&	\tbr{h}ini	&	\tbr{h}ini	&	turtle	\\
%						&	*ə\tbr{p}at			&	aat					&							&	aat	\\
						&	*ha\tbr{p}əjis	&	ili					&	ili					&	sick, pain	\\
						&	*ha\tbr{p}əjəs	&	\tbr{h}edaʔ	&	\tbr{h}eda	&	fever	(irr. *j > \it{d}, irr. *ə > \it{a})\\
						&	*qa\tbr{p}əju		&							&	ilu-n				&	gall(bladder)	\\
						&	*də\tbr{p}a			&							&	dea					&	fathom, armspan	\\
%						&	*ha\tbr{p}uy		&	ai					&	ai					&	fire	\\
						&	*hi\tbr{p}aR		&	i\tbr{h}aʔ	&	i\tbr{h}a		&	woman’s female cross-cousin	\\
%						&	*mani\tbr{p}is	&	mni\tbr{h}is&	ni\tbr{h}is	&	thin	\\
						&	*ma-\tbr{p}utiq	&	mutiʔ				&	mutiʔ				&	white		\\
						&	*sa-ŋa-\tbr{p}uluq	&	smulu		&	hŋulu		&	ten	\\
						&	*sa\tbr{p}a			&	saa	&	saa		&	what	\\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}

Within \tsc{Timor-Babar}, \ili{Helong} has a mix of conservative
and innovative features which set
it apart from other languages of this subgroup.
Three conservative retentions are *b = \it{b} /{\#\gap},
*d = \it{d} /{\#\gap}, and *ŋ = \it{ŋ}.
Examples are given in \qf{Hel-b}--\qf{Hel-ng}.
(Word medially *b > \it{h},\footnote{
		There are also a handful of cases of word-initial *b > \it{h} in \ili{Helong}.
		Of 32 instances of P(CE)MP *b in \cite{Edwards2018b} seven have *b > \it{h}:
		*bayu > \it{hai} `pound', *baRəq > \it{hale} `swell', *baŋun > \it{haŋu} `wake up',
		*busuk > \it{hosoʔ} `rotten', *buni > \it{huniʔ}, PCEMP **buu > \it{huu} `blow',
		and PCMP **belas > \it{khelaʔ} `machete'.}
and *d > \it{l}, see below.)

The only other \tsc{Timor-Babar}
languages which regularly retain *ŋ unchanged
are \ili{Hresuk} and \ili{Raklungu} (\srf{04-05}).
The only other language with *d = \it{d} is \ili{Makuva} (\srf{04-06-01})
and the only other language which unambiguously
has *b- = \it{b} is \ili{Ilih} East \ili{Damer} (\srf{04-06-03}).

{\wg\st{6pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lll>{\raggedright\arraybackslash}X}
\lsptoprule
						&	PMP						&	\ili{Funai}		&	\ili{Semau}	&	local gl.		\\ \midrule	\hhline{~}
%\endfirsthead
						%&	PMP						&	\ili{Funai}		&	\ili{Semau}	&	local gl.		\\ \midrule
\endhead
\tbx{Hel-b}	&	*\tbr{b}abuy	&	\tbr{b}ahi		&	\tbr{b}ahi	&	pig	\\
						&	*\tbr{b}aki		&	\tbr{b}akiʔ		&	\tbr{b}aki	&	mother's brother	\\
						&	*\tbr{b}asəq	&	\tbr{b}ase		&	\tbr{b}ase	&	wash	\\
						&	*\tbr{b}atu		&	\tbr{b}atu		&	\tbr{b}atu	&	stone, rock	\\
						&	*\tbr{b}aqbaq	&	\tbr{b}ahaʔ		&	\tbr{b}aha	&	mouth	\\
						&	*\tbr{b}ituqən&	\tbr{b}duun		&	duun				&	star	\\ \midrule\hhline{~}
\tbx{Hel-d}	&	*\tbr{d}akəp	&								&	\tbr{d}aken		&	catch, arrest	\\
						&	*\tbr{d}akin	&	\tbr{d}aki		&								&	dirty	(body, clothes)	\\
						&	*(a)da[nŋ]i		&								&	\tbr{d}aŋi		&	near	\\
						&	*\tbr{d}aRaq	&	\tbr{d}ala		&	\tbr{d}ala		&	blood	\\
						&	*\tbr{d}aRəq	&	k\tbr{d}ale		&	\tbr{d}ale		&	earth	\\
						&	*\tbr{d}əpdəp	&	\mc{2}{c}{k\tbr{d}ela}	&	\it{Erythrina} tree (unspecified variety of \ili{Helong} from \citealt[332]{Jonker1908})	\\
						&	*\tbr{d}iRi		&	\tbr{d}ili		&	\tbr{d}ili		&	stand	\\
						&	*\tbr{d}iRus	&	\tbr{d}iu			&	\tbr{d}iuʔ		&	bathe	\\
						&	*\tbr{d}uRi		&								&	\tbr{d}uliʔ		&	thorn	\\
						&	*\tbr{d}uha		&	\tbr{d}ua			&	\tbr{d}ua			&	two	\\
						&	*\tbr{d}uyuŋ	&	\mc{2}{c}{\tbr{d}uiŋ}					&	dugong (unspecified variety of \ili{Helong} from \citealt[82]{Jonker1908})	\\ \midrule\hhline{~}
\tbx{Hel-ng}&	*\tbr{ŋ}ajan	&	\tbr{ŋ}alaʔ		&	\tbr{ŋ}ala		& name		\\
						&	*sa-\tbr{ŋ}a-Ratus	&	\tbr{ŋ}atus		&	\tbr{ŋ}atus		& hundred		\\
						&	*\tbr{ŋ}ilu		&	\tbr{ŋ}iluʔ,	&	\tbr{ŋ}ilu		& sour		\\ \hhline{~}
						&								& \tbr{ŋ}iduʔ		&								& \hp{\,}\\
						&	*bəR\tbr{ŋ}i	&	bi\tbr{ŋ}in	dua	&	bi\tbr{ŋ}in	& \ili{Funai} `day after tomorrow', \ili{Semau} `day', PMP `night'		\\
						&	*ba\tbr{ŋ}un	&	ha\tbr{ŋ}u		&	ha\tbr{ŋ}u		& wake up		\\
						&	*na\tbr{ŋ}uy	&	na\tbr{ŋ}en		&	na\tbr{ŋ}en		& swim		\\
						&	*suma\tbr{ŋ}əd&	sma\tbr{ŋ}in	&	hma\tbr{ŋ}in	& forehead, spirit, soul		\\
						&	*ta\tbr{ŋ}ila	&	k\tbr{ŋ}ilaʔ	&	h\tbr{ŋ}ila		& ear		\\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}

One sound change which is unique to \ili{Helong} is a
partial merger of *l/*R/*j/*z/*d.
Word initially and medially *l/*R/*j/*z usually merge as \it{l},
with a few cases of *R > {\0} and a few of *z/*d/*j > \it{d}.
Word initially, *d = \it{d}, as illustrated in \qf{Hel-d} above,
but medially *d > \it{l}, thus merging with *l/*R/*j/*z.
Examples are given in \qf{Hel-l}--\qf{Hel-d-med}.

{\wg\st{6pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lll>{\raggedright\arraybackslash}X}
\lsptoprule
						&	PMP									&	\ili{Funai}					&	\ili{Semau}					&	local gl.		\\ \midrule	\hhline{~}
%\endfirsthead
						%&	PMP}							&	\ili{Funai}					&	\ili{Semau}					&	local gl.		\\
\endhead
\tbx{Hel-l}	&	*\tbr{l}akaw				&	\tbr{l}ako		&	\tbr{l}ako		&	go	\\
						&	*\tbr{l}aki					&	\tbr{l}ahi		&	\tbr{l}ahi		&	king	\\
%						&	*\tbr{l}aləj				&	\tbr{l}ales		&	h\tbr{l}ales	&	fly (n.)	\\
%						&	*\tbr{l}aRiw				&	\tbr{l}ali		&	\tbr{l}ali		&	run, flee	\\
						&	*\tbr{l}awaq				&	li\tbr{l}aun	&	\cg{(buskao)}&	spider	\\
%						&	**\tbr{l}əmba				&	\tbr{l}epa		&	\tbr{l}epa		&	carry on shoulder with pole (One person with something on each end, or two or more people with one heavy thing in the middle)	\\
						&	**\tbr{l}əmba				&	\tbr{l}epa		&	\tbr{l}epa		&	carry on shoulder with pole \\
%						&	*\tbr{l}i\tbr{l}in	&	\tbr{l}i\tbr{l}in	&						&	candle, wax (possible \ili{Malay} loan)	\\
						&	*\tbr{l}ima					&	\tbr{l}ima		&	\tbr{l}ima		&	five	\\
						&	*ba\tbr{l}abaw			&	b\tbr{l}aho		&	b\tbr{l}aho		&	mouse, rat	\\
						&	*bu\tbr{l}an				&	bu\tbr{l}an		&	bu\tbr{l}an		&	moon, month	\\
						&	*bu\tbr{l}iR				&	bu\tbr{l}in		&	bu\tbr{l}in		&	grainhead	\\
						&	*bu\tbr{l}u					&	bu\tbr{l}u		&	bu\tbr{l}u		&	body hair	\\
%						&	*da\tbr{l}əm				&	da\tbr{l}en		&	da\tbr{l}e		&	inside	\\
						&	*ha\tbr{l}as				&	a\tbr{l}as		&	a\tbr{l}as		&	forest	\\
						&	*ku\tbr{l}it				&	k\tbr{l}uit		&	\cg{(bulaŋ)}	&	skin (metathesis of *ul > \it{lu})	\\
						&	*ku\tbr{l}uR				&	ku\tbr{l}uʔ		&								&	breadfruit	\\
%						&	*la\tbr{l}əj				&	la\tbr{l}es		&	hla\tbr{l}es	&	fly (n.)	\\
%						&	*\tbr{l}əsuŋ				&	isuŋ					&								&	mortar	\\
						&	**ma\tbr{l}ip				&	ma\tbr{l}i		&	ma\tbr{l}i		&	laugh	\\
%						&	**mə\tbr{l}ayu			&	b\tbr{l}ai		&	b\tbr{l}ai		&	foreign	\\
						&	*ŋi\tbr{l}u					&	ŋi\tbr{l}uʔ,	&	ŋi\tbr{l}u		&	sour	\\ \hhline{~}
						&	\vp{.}							&	ŋi\tbr{d}uʔ		&	\vp{.}				&	\vp{.}		\\
						&	*qahə\tbr{l}u				&	a\tbr{l}u			&	a\tbr{l}u			&	pestle	\\
%						&	*qa\tbr{l}əjaw			&	\tbr{l}elo		&	\tbr{l}elo		&	\ili{Funai} = `sun, sky', \ili{Semau} `sun'	\\
						&	*qatimə\tbr{l}a			&	kme\tbr{l}aʔ	&								&	flea (\citet[614]{Jonker1908} has unspecified \ili{Helong} \it{hmela})\\
						&	*qatə\tbr{l}uR			&	ti\tbr{l}un		&	ti\tbr{l}un		&	egg	\\
%						&	*qu\tbr{l}əj				&	u\tbr{l}as		&	u\tbr{l}as		&	\ili{Funai} = `snake', \ili{Semau} = `maggots'	\\
%						&	*sa-ŋa-pu\tbr{l}uq	&	smu\tbr{l}u		&	hŋu\tbr{l}u		&	ten	\\
						&	*ta\tbr{l}ih					&	ta\tbr{l}i		&	ta\tbr{l}i	&	rope	\\
%						&	*taŋi\tbr{l}a				&	kŋi\tbr{l}aʔ	&	hŋi\tbr{l}a		&	ear	\\
%						&	*tu\tbr{l}aq				&	k\tbr{l}utiʔ	&	k\tbr{l}uti		&	gewang palm	(consonant metathesis)\\
						&	*tə\tbr{l}u					&	ti\tbr{l}u		&	ti\tbr{l}u		&	three	\\
%						&	*wa\tbr{l}u					&	fa\tbr{l}u		&	pa\tbr{l}u		&	eight	\\
%						&	*za\tbr{l}an				&	la\tbr{l}an		&	la\tbr{l}an		&	path, way	\\
\midrule\hhline{~}
\tbx{Hel-R}	&	*\tbr{R}abiqi				&	s\tbr{l}ahin	&	k\tbr{l}ahin	&	yesterday	\\
						&	*\tbr{R}ibu					&	\tbr{l}ihu		&	\tbr{l}ihu		&	thousand	\\
						&	*baqə\tbr{R}u				&	ba\tbr{l}u		&	ba\tbr{l}u		&	new	\\
						&	*ba\tbr{R}əq				&	ha\tbr{l}e		&	ha\tbr{l}e		&	swell (irr. *b > \it{h} /{\#\gap}	\\
						&	*bə\tbr{R}ay				&	be\tbr{l}e		&	be\tbr{l}e		&	give	\\
						&	*da\tbr{R}aq				&	da\tbr{l}a		&	da\tbr{l}a		&	blood	\\
						&	*da\tbr{R}əq				&	kda\tbr{l}e		&	da\tbr{l}e		&	earth	\\
						&	*di\tbr{R}i					&	di\tbr{l}i		&	di\tbr{l}i		&	stand	\\
						&	*du\tbr{R}i					&								&	du\tbr{l}iʔ		&	thorn	\\
						&	*ha\tbr{R}əzan			&								&	e\tbr{l}an		& ladder, stairs		\\
						&	*ka\tbr{R}us				&	ka\tbr{l}o		&	ka\tbr{l}o		&	scratch	\\
						&	*la\tbr{R}iw				&	la\tbr{l}i		&	la\tbr{l}i		&	run, flee	\\
						&	*ma\tbr{R}uqanay		&	b\tbr{l}anen	&	b\tbr{l}anen	&	woman's brother	\\
						&	*pə\tbr{R}əq				&	\mc{2}{c}{he\tbr{l}e}					&	squeeze out (unspecified variety of \ili{Helong} from \citealt[169]{Jonker1908})	\\
						&	*qasi\tbr{R}a				&	si\tbr{l}a		&	si\tbr{l}a		&	salt	\\
						&	*ta\tbr{R}a					&	ta\tbr{l}a		&	ta\tbr{l}a		&	answer	\\
						&	*ta\tbr{R}um				&	\mc{2}{c}{ta\tbr{l}uŋ}				&	indigo plant	(unspecified variety of \ili{Helong} from \citealt[770]{Heyne1950})	\\
						&	*tə\tbr{R}as				&								&	te\tbr{l}as		&	beam, strong; structurally sound, not corrupted by mould or rot (PMP `heartwood')	\\
						&	**tubu\tbr{R}a			&	thu\tbr{l}aʔ	&	hu\tbr{l}a		&	lips (**tubuRa = `spit')	\\
						&	*u\tbr{R}at					&	u\tbr{l}at, 	&	u\tbr{l}at		&	\hp{\,}\\ \hhline{~}
						&	\hp{\,}							&	u\tbr{d}at		&	\hp{\,}				&	\multirow{-2}{*}{vein, fortune (\ili{Funai} /d/ → [d] {\tl} [r])}	\\ \midrule\hhline{~}
						&	*\tbr{R}umaq				&	uma						&	uma						&	house	\\
						&	*bə\tbr{R}as				&	beas					&	beas					&	seed, uncooked rice	\\
						&	*haba\tbr{R}at			&								&	haat					&	west monsoon	\\
						&	*ma-i\tbr{R}aq			&	meaʔ					&	mea						&	red, brown	\\
						& *ti\tbr{R}əm				&	\mc{2}{c}{tiaŋ}								&	oyster (unspecified \ili{Helong} from \citealt[629]{Jonker1908})	\\
						& *pa\tbr{R}ih				&	\mc{2}{c}{ai}									&	stingray (unspecified \ili{Helong} from \citealt[153]{Jonker1908})	\\
						&	*za\tbr{R}um				&	lauŋ					&	lauŋ					&	needle	\\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}

{\wg\st{6pt}\begin{xltabular}[l]{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}>{\hspace{-\tabcolsep}\enspace}lll>{\raggedright\arraybackslash}X}
\lsptoprule
						&	PMP									&	\ili{Funai}					&	\ili{Semau}					&	local gl.		\\ \midrule	\hhline{~}
%\endfirsthead
						%&	PMP}							&	\ili{Funai}					&	\ili{Semau}					&	local gl.		\\
\endhead
\tbx{Hel-z}	&	*\tbr{z}alan				&	\tbr{l}alan		&	\tbr{l}alan		& path, way		\\
						&	*\tbr{z}aRum				&	\tbr{l}auŋ		&	\tbr{l}auŋ		& needle		\\
						&	*\tbr{z}aqat				& \tbb{d}aat		& \tbb{d}aat		& bad, evil, rotten		\\
						&	*\tbr{z}aŋkal				&								&	\tbb{d}aŋa		& span of hand, inch		\\
						&	*haRə\tbr{z}an			&								&	e\tbr{l}an		& ladder, stairs		\\
						&	*qu\tbr{z}an				&	u\tbr{l}an		&	u\tbr{l}an		& rain		\\
						&	*tu\tbr{z}uq				&								&	tu\tbr{l}u		& point, show, designate		\\
						&	**qa\tbr{z}ay				&	a\tbb{s}aʔ		&	a\tbb{s}aʔ		& chin	(irr. *ay > \it{a})\\ \midrule\hhline{~}
\tbx{Hel-j}	&	*hapə\tbr{j}is			&	i\tbr{l}i			&	i\tbr{l}i			&	sick, pain	\\
						&	*hua\tbr{j}i				&	fa\tbr{l}in		&	pa\tbr{l}in		&	same sex younger sibling	\\
						&	*i\tbr{j}uŋ					&	i\tbr{l}uʔ		&	i\tbr{l}u			&	nose	\\
						&	*ŋa\tbr{j}an				&	ŋa\tbr{l}aʔ		&	ŋa\tbr{l}a		&	name	\\
						&	*pa\tbr{j}ay				&	a\tbr{l}e			&	a\tbr{l}e			&	rice plant	\\
						&	*pi\tbr{j}a					&	i\tbr{l}a			&	i\tbr{l}a			&	how many?	\\
						&	*qalə\tbr{j}aw			&	le\tbr{l}o		&	le\tbr{l}o		&	\ili{Funai} = `sun, sky', \ili{Semau} `sun'	\\
						&	*bu\tbr{j}əq				&	bu\tbb{d}at		&	bu\tbr{l}at		&	foam (\ili{Funai} possibly has variant **\it{bulat}	\\
						&	*hapə\tbr{j}əs			&	he\tbb{d}aʔ		&	he\tbb{d}a		& fever	(irr. *ə > \it{a})\\  \midrule\hhline{~}
\tbx{Hel-d-med}	&	*ma-\tbr{d}iŋdiŋ&	b\tbr{l}iŋin	&	b\tbr{l}iŋin	&	cold, cool	\\
						&	*qu\tbr{d}ip				&	nu\tbr{l}iʔ		&	nu\tbr{l}iŋ		&	live (v.)	\\
						&	**sə\tbr{d}u(t)			&	si\tbr{l}u		&								&	weaving sword	(cf. E. \ili{Tetun} \it{soru}, Proto-Rote-\ili{Meto} \it{seru}, Hawu \it{pehəɗu})\\
						&	*tu\tbr{d}iq				&	\mc{2}{c}{tu\tbr{l}is}				&	pigeon pea (unspecified variety of \ili{Helong} from \citealt[82]{Jonker1908})	\\
						&	**baŋku\tbr{d}u			&	\mc{2}{c}{(p)ku\tbb{d}u}			&	k.o. tree (unspecified variety of \ili{Helong} from \citealt[347]{Jonker1908})	\\
						&	*mu\tbr{d}a					&	mu\tbb{d}aʔ		&	mu\tbb{d}a		&	young	\\
						&	*ma-u\tbr{d}əhi			&								&	mu\tbb{d}i		&	follow	\\
						&	*ŋa\tbr{d}a					&								&	ŋa\tbb{d}a 		& look up, come before \\
\lspbottomrule
\end{xltabular}\addtocounter{table}{-1}}

Word finally, *l/*r/*d are usually lost
(*z does not occur finally), but *j > \it{s}, though
only three examples have been identified.
Final *j > \it{s} is not only unique within \tsc{Timor-Babar}
it is almost unique within the entire region of eastern Indonesia.
Only some members of \tsc{Seram-Tanimbar-Bomberai} (\srf{11-03}, \srf{11-05})
as well as all members of SHWNG have *j > \it{s}.
Examples of final *j in \ili{Helong} are given in \qf{Hel-j-fin}.

\noindent{\wg\st{6pt}\begin{tabularx}
{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}
>{\hspace{-\tabcolsep}\enspace}lll
>{\raggedright\arraybackslash}X}
\lsptoprule
						&	PMP			&	\ili{Funai}					&	\ili{Semau}					&	local gl.		\\ \midrule
\tbx{Hel-j-fin}
						&	*laləj	&	lale\tbr{s}		&	hlale\tbr{s}	&	fly (n.)	\\
						&	*quləj	&	ula\tbr{s}		&	ula\tbr{s}		&	\ili{Funai} = `snake', \ili{Semau} = `maggots'	\\
						&	*kunij	&	kuni\tbr{s}		&	kuni\tbr{s}		&	yellow, green, blue (PMP `turmeric') 	\\
						&	*bukij	&								&	huin					&	wild, untamed (PMP `mountain, forest' cf. Proto-Rote-\ili{Meto} **fui with same semantic shift)	\\
						&	*bəkəlaj&								&	bela					&	open eyes (PMP `spread out, unroll (mats, etc.), open out, unfold (as the hand); wide')	\\
\lspbottomrule
\end{tabularx}\mbox{}\\}

Another sound change which is mostly unique
to \ili{Helong} is *w > **f, though this change is partially
shared with \tsc{Rote-Meto} which has a three-way
split of *w, of which one out come is **f (\srf{04-02}; \citealp[82ff]{Edwards2021}).
\ili{Semau} and \ili{Bolok} \ili{Helong} have subsequent **f > \it{p}.
Examples are given in \qf{Hel-w}.

\noindent{\wg\st{6pt}\begin{tabularx}
{\linewidth}{>{\hspace{-\tabcolsep}}l<{\hspace{-\tabcolsep}}
>{\hspace{-\tabcolsep}\enspace}lll
>{\raggedright\arraybackslash}X}
\lsptoprule
						&	PMP						&	\ili{Funai}				&	\ili{Semau}				&	local gl.		\\ \midrule
\tbx{Hel-w}	&	*a\tbr{w}aŋ		&	a\tbr{f}on	&	a\tbr{p}an	&	sky	\\
						&	*ha\tbr{w}ak	&	a\tbr{f}aʔ	&	a\tbr{p}a		&	body, self (PMP `waist')	\\
						&	*qasa\tbr{w}a	&	sa\tbr{f}a	&	sa\tbr{p}a	&	spouse	\\
						&	*h\tbr{ua}ji	&	\tbr{f}alin	&	\tbr{p}alin	&	same sex ySib	\\
						&	*\tbr{w}alu		&	\tbr{f}alu	&	\tbr{p}alu	&	eight	\\
						&	*qatay > **\tbr{w}atay	&	\tbr{f}aten	&	\tbr{p}ate	&	liver	(cf. \ili{Sika} \it{ʔβate} `liver')\\
						&	*\tbr{w}ahiyəR&	ui					&	ui					&	water	\\
						&	*si\tbr{w}a		&	si\tbb{p}a	&	si\tbr{p}a	&	nine (irr. *w > \it{p} in \ili{Funai})	\\
						&								&	kla\tbr{f}a	&	kla\tbr{p}a	&	garden, field	\\
						&								&	\tbr{f}alun	&	\tbr{p}alun	&	river	\\
\lspbottomrule
\end{tabularx}\mbox{}\\}

In the lexicon \ili{Helong} also has a mix of conservative
and innovative features.
Unique semantic shifts include *buaq ‘fruit’ > \it{bua} ‘banana’
(\it{isin} = ‘fruit’), and *asu ‘dog’ > \it{asu} ‘cow’ (\it{ŋoot} = ‘dog’).
One semantically conservative word is \it{duliʔ} ‘thorn’ from PMP *duRi ‘thorn’.
Most languages of the \tsc{Timor-Babar} subgroup -- like most
languages of linguistic Wallacea -- have undergone semantic
shift of *duRi to mean ‘bone’.
(\ili{Helong} has \it{seen} for ‘bone’.) Similarly,
\ili{Helong} \it{sila} ‘salt’ appears to be the only retention of PMP
*qasiRa ‘salt’ in \tsc{Timor-Babar}.
Other \tsc{Timor-Babar} languages have reflexes of *ma-qasin
or *qasin for ‘salt’.