\scp{\tsc{East Sumbawa}}{03-01}
The exact number of languages/dialects spoken in the eastern
part of \ili{Sumbawa} Island (corresponding to the current Indonesian
regencies of \ili{Bima} and Dompu) remains to be properly investigated.
At l\ili{east} three languages are identifiable: \ili{Bima}, \ili{Kolo}, and \ili{Sanggar}.
The largest and most well described is \ili{Bima} [bhp] (a.k.a.
\ili{Mbojo}).\footnote{
		In May 2015, four \ili{Mbojo} speakers (both male
		and female) from East \ili{Sumbawa} met briefly with Grimes
		and explained that their language was different from \ili{Bima} spoken in the city.
		It may be both a different dialect
		and an alternate name for the language.
		The exact basis of the difference is unclear,
		but the idea of “\ili{Mbojo}” being an alternate name,
		and also a dialect/variety of \ili{Bima} needs to be pursued further.}
\ili{Sanggar} is spoken on the eastern part of the \ili{Sanggar} peninsula.
At the time of writing,
it has not been assigned an ISO 693-3 code,
but the lexical data in \citet{Zollinger1850} shows that it is
clearly a separate language from \ili{Bima}.\footnote{
		Thanks to Harald Hammarström for bringing the existence of \ili{Sanggar}
		to our attention.} 

In addition to these two languages,
\citet[3--4]{Jonker1896} reports the existence of two “dialects” of
\ili{Bima}: \ili{Kolo} (spoken in the eponymous village on the north-eastern
part of the bay of \ili{Bima} \citep{PartamiKaryawan1995}, and \ili{Toloweri}.
Although available \ili{Kolo} data is limited,
it has differences in sound changes,
functors, basic vocabulary, and morphology compared with \ili{Bima}.
Thus, we separate it off as a separate language.\footnote{
		\citet[4]{Jonker1896}
		writes: “\ili{Kolo} can be considered a dialect of \ili{Bima}.” However,
		Jonker uses the term “dialect” for speech varieties which are
		separate languages, though more closely related to one another
		than to other neighbouring languages.
		Thus, for instance, all the different languages of Rote Island
		(\srf{04-02}) are labelled “dialects” by Jonker.}
Jonker reports
that “The language of Tolo-Weri corresponds completely to that
of \ili{Kolo} with regard to sound changes,
and to a great extent also with regard to vocabulary”.\footnote{
“De
taal van Tòlo-Weri komt volkomen overeen met die van Kòlo ten
opzichte der klankverschijnselen en grootendeels ook wat betreft
den woordenschat,” \citep[4]{Jonker1896}.}
However,
almost no \ili{Toloweri} data is available to make an independent
assessment of its relationship with \ili{Kolo} or \ili{Bima}.

There are a number of changes that are shared between \ili{Bima},
\ili{Kolo}, and \ili{Sanggar} which allow them to be placed within an \tsc{East
Sumbawa} group: *R > {\0} in most words, *-ay > \it{e,}
and *-aw > \it{o}.
All of these sound changes also occur in \ili{Komodo} (\srf{03-01-01}),
as well as certain other \tsc{\ili{Bima}-Lembata} languages,
thus considerably weakening the evidence for \tsc{East Sumbawa}
as a distinct subgroup within \tsc{\ili{Bima}-Lembata}.

Among the \tsc{East Sumbawa} languages,
\ili{Bima} shows a large number of unconditioned splits,
most of which are summarised by \citet[75--83]{Blust2008}.
The following unconditioned splits occur in \ili{Bima}: *p > \it{f,
p}; *t > \it{d,
t}; *k > \it{h,
k}; initial *b- > \it{ɓ,
b, mb, w}; initial *d- > \it{r,
ɗ}; *j/*l > \it{r,
l}; *z > \it{r,
ɗ}; *s > \it{s,
ʧ}; *R > \it{r,} {\0};
and *w > \it{w,} {\0}.
Some of these splits have partial conditioning environments,
but exceptions to these environments also occur.
For the most part these splits are not shared with \ili{Sanggar} or \ili{Kolo}.
The changes *t > \it{d,
t}, *l > \it{r, l},
and (perhaps) *p > \it{f,
p} and *s > \it{s,
ʧ} are shared with \tsc{\ili{Havu}-Dhao} (\srf{03-03-03}).
Similarly, there are some cases of medial *k > \it{h} shared
between \ili{Bima} and \ili{Komodo}.

There is also disagreement between \citet{Zollinger1850}
and more modern sources as to whether a certain word attests
a particular sound change in \ili{Bima}.
Four examples which appear unlikely to be due to error are the
following (words from \citet{Zollinger1850} are given first): *saŋan
> \it{ʧaŋa}, \it{saŋa} ‘branch’; *kutu > \it{kudu},
\it{hudu} ‘headlouse’, *siku > \it{ʧiku}, \it{ʧihu} ‘elbow’,
and *baqəRu > \it{wou}, \it{ɓou} ‘new’.
Whether these differences are due to dialectal differences or
have some other basis is currently unknown.

The reflexes of *ə are particularly diverse in \ili{Bima},
with the syllable position
and qualities of adjacent vowels
and consonants all playing a role in conditioning which reflexes occur.
\ili{Sanggar} usually has *ə > \it{e} while \ili{Kolo} usually has *ə > \it{o}
in available data, though exceptions occur.
Before a syllable containing \it{u} \ili{Kolo}
and \ili{Bima} have *ə > \it{u},
and \ili{Sanggar} has *ə > \it{e}.
Examples are given in \trf{03-05}.

\begin{table}[p]
	\centering\tcp{\tsc{East Sumbawa} *ə /{\gap}(C)u}{03-05}
	\st{4.25pt}\begin{tabular}{llllllll}\lsptoprule
*gloss&egg&sugarcane&three&new&PP/CC&fart&gall\\
PMP&*qat\tbr{ə}luR&*t\tbr{ə}buh&*t\tbr{ə}lu&*baq\tbr{ə}Ru&*\tbr{ə}mpu&**p\tbr{ə}su&*qap\tbr{ə}ju\\ \midrule
\ili{Sanggar}&{\hr\hp{qa}}t\tbr{e}lu&{\hr}t\tbr{e}wu&{\hr}t\tbr{e}lu&{\hr}w\tbr{e}u&&{\hr\hr}p\tbr{e}su&\\
\ili{Kolo}&{\hr\hp{qa}}t\tbr{o}lu&{\hr}t\tbr{o}wu&&&{\hr}\tbr{o}mpu-na&&\\
\ili{Bima}&{\hr\hp{qa}}d\tbr{o}lu&{\hr}d\tbr{o}ɓu&{\hr}t\tbr{o}lu&{\hr}ɓ\tbr{o}u&{\hr}\tbr{o}mpu&{\hr\hr}p\tbr{o}ʧu&{\hr\hp{qa}}f\tbr{o}lu\\
		\lspbottomrule
	\end{tabular}
\end{table}

\ili{Bima} usually (though not universally) has *ə > \it{e,
i} before syllables containing \it{i},
including when this \it{i} is not a reflex of *i.
There are two examples of \ili{Kolo} *ə > \it{o} in this environment
and two of \ili{Sanggar} *ə > \it{e}.
Examples of *ə before syllables containing \it{i} in \ili{Bima} are
given in \trf{03-06}.

\begin{table}[p]
	\centering\tcp{\tsc{East Sumbawa} *ə /{\gap}(C)i}{03-06}
	\begin{tabular}{llllllll}\lsptoprule
*gloss	&	buy	&	low-tide	&	give	&	sneeze	&	full	&	stand	&	six	\\	
PMP	&	*b\tbr{ə}li	&	*ma-q\tbr{ə}ti	&	*b\tbr{ə}Ray	&	*b\tbr{ə}ñan	&	*p\tbr{ə}nuq	&	*k\tbr{ə}dəŋ	&	*\tbr{ə}nəm	\\	\midrule
\ili{Sanggar}	&	{\hr}w\tbr{e}li	&	{\hr}m\tbr{e}ti	&	\<b\tbr{è}kè\>	&		&	\cg{(nisi)}	&	\cg{(tendi)}	&		\\	
\ili{Kolo}	&	{\hr}w\tbr{o}li	&	{\hr}m\tbr{o}ti	&	{\hr}w\tbr{e}ya	&		&		&	{\hr}h\tbr{o}ru	&	{\hr}\tbr{o}nu	\\	
\ili{Bima}	&	{\hr}w\tbr{e}li	&	{\hr}m\tbr{o}ti	&	{\hr}mb\tbr{e}i	&	{\hr}ɓ\tbr{e}ni	&	{\hr}ɓ\tbr{i}ni	&	{\hr}k\tbr{i}di	&	{\hr}\tbr{i}ni	\\	
		\lspbottomrule
	\end{tabular}
\end{table}

Before a syllable containing \it{a},
the reflexes are mixed,
with no clear conditioning environments,
except the presence of labial *p in *dəpa ‘fathom’
and *əpat `four' probably conditioning *ə > \it{u}.
Examples are given in \trf{03-07}.

\begin{table}[p]
	\centering\tcp{\tsc{East Sumbawa} *ə /{\gap}(C)a}{03-07}
	\st{5.5pt}\begin{tabular}{llllllll}\lsptoprule
*gloss&alone&one&hear&fathom&four&heartwood&heavy\\
PMP&*ma-\tbr{ə}sa&*\tbr{ə}sa&*d\tbr{ə}ŋəR&*d\tbr{ə}pa&*\tbr{ə}pat&*t\tbr{ə}Ras&*b\tbr{ə}Rəqat\\ \midrule
\ili{Sanggar}&{\hr}m\tbr{e}sa&&{\hr}pal\tbr{i}ŋa&{\hr}nd\tbr{u}pa&&\cg{(‹ka-ekh›)}&\cg{(dodo)}\\
\ili{Kolo}&{\hr}m\tbr{e}sa&{\hr}\tbr{o}sa&{\hr}pal\tbr{i}ŋa&&{\hr}\tbr{u}pa&‹kat\tbr{ò}o›&\\
\ili{Bima}&{\hr}k\tbr{e}se&&{\hr}r\tbr{i}ŋa&{\hr}nd\tbr{u}pa&{\hr}\tbr{u}pa&{\hr}t\tbr{e}ra&{\hr}b\tbr{a}ra\\
		\lspbottomrule
	\end{tabular}
\end{table}

In final syllables, \ili{Sanggar} usually has *ə > \it{o,} \ili{Kolo} usually
has *ə > \it{u,}
and \ili{Bima} usually has *ə > \it{i},
\it{a} though there are exceptions in all three languages.
Examples are given in \trf{03-08}.

\begin{table}[p]\centering\tcp{\tsc{East Sumbawa} *ə /σ\und{σ}{\#}}{03-08}
	\begin{threeparttable}\st{3pt}
	\begin{tabular}{lllll@{}ll@{}l}\lsptoprule
*gloss&stand&six&vine&inside&hear&wet&black\\
PMP&*kəd\tbr{ə}ŋ&*ən\tbr{ə}m&*waR\tbr{ə}j\su{†}&*dal\tbr{ə}m\su{‡}&*dəŋ\tbr{ə}R&*bas\tbr{ə}q&*ma-qit\tbr{ə}m\\	\midrule
\ili{Sanggar}&\cg{(tendi)}&&\cg{(ika)}&‹dal\tbr{o}kh›, lal\tbr{o}k&{\hr}paliŋ\tbr{a}&\cg{(ŋgesak)}&{\hr}mut\tbr{o}k\\
\ili{Kolo}&{\hr}hor\tbr{u}&{\hr}on\tbr{u}&{\hr}a\tbr{u}&‹kaḍal\tbr{u}›&{\hr}paliŋ\tbr{a}&&{\hr}kamet\tbr{e}\\
\ili{Bima}&{\hr}kid\tbr{i}&{\hr}in\tbr{i}&{\hr}a\tbr{i}&\cg{(ɗei)}&{\hr}riŋ\tbr{a}&{\hr}mbeʧ\tbr{a}&{\hr}meʔ\tbr{e}\\
		\lspbottomrule
	\end{tabular}
		\begin{tablenotes}\tnsize
			\item[†]	{Reflexes here mean ‘rope’.}
			\item[‡]	{Reflexes here mean ‘deep’.}
		\end{tablenotes}
	\end{threeparttable}
\end{table}

