\chapter{Language list and data sources}\label{ch:LanLis}
This appendix lists the languages that appear in this book,
along with the sources of data for each language.
%\footnote{
		%This appendix can be made available as a spreadsheet for interested
		%readers by contacting the authors.}
It also lists the ISO 639-3
code, Glottocode \citep{Hammarstrom2023},
and geographical coordinates for each language.
For languages in our target region,
this list is comprehensive (based on current knowledge),
though it does not systematically list every variety of each
language where these are known to be dialects in the linguistic sense.
For a small number of languages in our target region,
no data is presented in this book -- usually because no data is available.
These languages are listed,
but no source is given.

In addition to being a (near comprehensive) list of languages
in our target region,
this appendix also lists languages outside our target region
for which data is used in this book,
even if it is only a data point in one of the many maps.\footnote{
		This appendix does not include languages which only occur in \frf{02-17}
		(complex numerals), \frf{13-01} (absence of /ŋ/),
		and/or \frf{02-16} (order of possessor).
		Those maps are composed from data in \citet{SchapperHammarstrom2013},
		\citet{DonohueDawson2013}, and \citet{Dryer2013b} respectively,
		supplemented by the sources in this appendix.
		The data sources for Figures \ref{fig:02-20}
		and \ref{fig:15-11} (banana terms) are given in appendix \ref{ch:BanTer}
		and are not usually repeated here.
		Likewise, the sources listed in \crf{ch:SurMar} which
		were only used for marsupial terms
		are not consistently repeated here.}

This appendix lists 559 languages/speech varieties.
Of these, 463 are Austronesian
and 96 are from various Papuan families.
335 of these languages/varieties are found within
our primary target
region within Linguistic Wallacea,
including 41 Papuan \tsc{Timor-Alor-Pantar} languages,
but not counting SHWNG,
\tsc{North Halmahera} and other Papuan languages of the Bird’s
Head and Cenderawasih Bay region.
294 of these languages/varieties are from the
Wallacean subgroups we have identified.
Data occurs in this book for 256 of these.
%This tally does not include the 3 \tsc{Taliabu} languages associated with \tsc{Celebic}.

We are extremely grateful to all those who have shared their
unpublished data with us
and given us permission to use it for comparative research.
Other published sources cited throughout the book are listed in the Bibliography.
The sections below are organised geographically and by subgroup.
%Sources saying “\citep{GrimesC1990}” indicates that the named
%investigator filled in \citet{GrimesC1990} by hand
%and sent a copy to Grimes for use in comparative studies.

The latitude and longitude values for languages within our target
region are updated where appropriate from those in Glottolog
\citep{Hammarstrom2023}.
The coordinates are roughly in the centre of the language region,
or on the island which is most prominent for the language/speech variety in question.
Coordinates for languages outside our target region are usually
unchanged from those in Glottolog.

“{-}{-}{-}” in the ISO column indicates that the language has
not yet been assigned an ISO 639-3 code,
with our judgment being that,
based on present evidence,
the language probably needs an individual code.
We do not directly address the issue where multiple varieties
share a common code,
such as several \tsc{Meto} cluster languages with different historical
sound correspondences, different morphophonemics,
and different vocabulary sharing the code [aoz].
Shared ISO codes are given in brackets.
Some languages sharing a code may need their own code, some do not.
We also do not distinguish extinct languages from those that are still spoken.
All those issues need to be addressed separately.

The ordering of languages within each subgroup is a compromise
between node and geography with nodes going from west to east and/or north to south.
Languages within a single node are grouped together,
and are similarly ordered from west to east and/or north to south.
The order of languages follows their order in the trees at the
end of each chapter.
All data in this appendix which is taken from 
\citet{SmitsVoorhoeve1992a,SmitsVoorhoeve1992b,
SmitsVoorhoeve1994,SmitsVoorhoeve1998} is abbreviated to “SV”.
\defcitealias{SmitsVoorhoeve1992a}{SV 1992a}
\defcitealias{SmitsVoorhoeve1992b}{SV 1992b}
\defcitealias{SmitsVoorhoeve1994}{SV 1994}
\defcitealias{SmitsVoorhoeve1998}{SV 1998}

\input{chapters/A-LanLis/A-01}
\input{chapters/A-LanLis/A-02}
\input{chapters/A-LanLis/A-03}
\input{chapters/A-LanLis/A-04}
\input{chapters/A-LanLis/A-05}
\input{chapters/A-LanLis/A-06}
\input{chapters/A-LanLis/A-07}
\input{chapters/A-LanLis/A-08}
\input{chapters/A-LanLis/A-09}
\input{chapters/A-LanLis/A-10}
\input{chapters/A-LanLis/A-11}
	\input{chapters/A-LanLis/A-11-01}
	\input{chapters/A-LanLis/A-11-02}
	\input{chapters/A-LanLis/A-11-03}
	\input{chapters/A-LanLis/A-11-04}
	\input{chapters/A-LanLis/A-11-05}
	\input{chapters/A-LanLis/A-11-06}
\input{chapters/A-LanLis/A-12}
\input{chapters/A-LanLis/A-13}
	\input{chapters/A-LanLis/A-13-01}
	\input{chapters/A-LanLis/A-13-02}
	\input{chapters/A-LanLis/A-13-03}