\chapter{Layers of added complexity}\label{ch:LayAdd}
In order to better understand the languages in the region as they are spoken today,
it helps to also understand the context in which they are spoken.
That is, the geography,
prehistory, and patterns of contact and trade.
It is also important to understand the history of the study of the languages,
and some of the conventions used in earlier eras to document these languages.

In looking to classify the Austronesian languages in the region,
one cannot wear Austronesian blinders.
When people speaking Austronesian languages first entered the region they did not come into a vacuum.
When they first began to arrive in this diverse region spanning almost 2,000 kilometres from west to
\ili{east}, just over 1,000 kilometres north to south,
and encompassing over 2,000 islands,
there were already other people on many of those islands speaking diverse languages that we subsume loosely
under the term Papuan.\footnote{
		Papuan does not define a single genealogical linguistic affiliation.
		It is a cover term of convenience used to refer to numerous unrelated language families in the greater
		New Guinea region that are not Austronesian,
		and not Australian Aboriginal
		(\citealp{Foley1986}, \citealp[1]{Klamer2017}).}

The speakers of Austronesian languages encountered new geography,
new climates, new flora
and fauna, new food possibilities, and new cultural ideas they had not previously had ways of talking
about, nor had they had any need for them.
\ili{As} they migrated, traded,
became bilingual and intermarried with other groups,
they encountered and accommodated new kin terms
and social categories that were deemed useful.
Different groups in different islands had significantly different patterns of contact and trade.
So, for example, some terms
and patterns found along the southern group of islands,
are not found in the northern islands, and vice versa.
And there are more localised patterns as well.
Some languages adopted new words.
Closely related neighbouring languages did not adopt those same words.
Some new words became conventionalised across whole clusters of languages; some only in individual languages.
Examples illustrating these phenomena are given in this and subsequent chapters.

Some early sources documenting languages in the region
were written by people with some training in (European or classical) languages.
Others were compiled by people with no special training.
So it should not be surprising that there is great variation in the notation conventions,
accuracy and reliability of the early documents,
as well as in the details.
Some of the early notational conventions were fairly widespread during a certain time period,
but not necessarily used consistently.
So, there are many
and various ways in the region to represent glottal stops,
phonetically long vowels, vowel systems that have more than five vowels,
word stress, ``funny'' consonant sounds, allophones, word breaks, etc.
Or these sounds were not represented at all by some people documenting the languages.
Some investigators used three different ways to indicate vowel length within a single language.
Some were consistent in their different practices for marking different long vowels,
and some were not.
Some body parts and kin terms are given in first person,
second person or third person forms.
Some investigators give just a noun or verb root,
with an indication that there is an inflection paradigm,
but without providing such a paradigm.
So we have some roots,
but not necessarily in the form of a full word that would be recognised by speakers of the languages.
Similarly for verbs, some investigators provide verbs with their inflections,
and others cite only the root,
without the information necessary to have a citation form,
and without a full paradigm for any verb.
This is complicated in that languages in many localised regions have morphophonemic changes in their
verbal paradigms, and the form provided may not directly reflect the normal historical sound correspondence for that language,
but rather reflects the result of a subsequent morphophonemic change.
European investigators often glossed local flora
and fauna with names of similar European flora
and fauna, and (in the classical tradition) cited verbs as infinitives which were actually inflected
finite verbs in the local languages.

All of these issues add layers of complexity to the task of unravelling the prehistory
and classification of languages in the region.
In \srf{02-01} we look at what is known about the geography,
climate, and flora and fauna in relation to the early prehistory of humans in the region,
providing a backdrop to the later arrival of the Austronesian speaking peoples.
In \srf{02-02} we look at evidence for various kinds of on-going contact,
migrations, and trade that continued after the arrival of the Austronesian speaking peoples,
showing that there were many external factors that continued to impact the development
and ongoing diversification of languages in the region.
In \srf{02-03} we look at various kinds of evidence found in the languages themselves for the kinds of
contact-induced change described in the previous two sections.
And in \srf{02-04} we describe various issues in understanding
and unravelling the transcription of the languages necessary for careful comparative work.
Some of these issues have been ignored or misunderstood by comparative linguists,
and this has contributed to confusion with regard to the languages under investigation in this book.

\input{chapters/02-LayAdd/02-01}
\input{chapters/02-LayAdd/02-02}
\input{chapters/02-LayAdd/02-03}
	\input{chapters/02-LayAdd/02-03-01}
	\input{chapters/02-LayAdd/02-03-02}
		\input{chapters/02-LayAdd/02-03-02-01}
		\input{chapters/02-LayAdd/02-03-02-02}
		\input{chapters/02-LayAdd/02-03-02-03}
		\input{chapters/02-LayAdd/02-03-02-04}
		\input{chapters/02-LayAdd/02-03-02-05}
\input{chapters/02-LayAdd/02-04}
	\input{chapters/02-LayAdd/02-04-01}
	\input{chapters/02-LayAdd/02-04-02}
	\input{chapters/02-LayAdd/02-04-03}
	\input{chapters/02-LayAdd/02-04-04}
	\input{chapters/02-LayAdd/02-04-05}
	\input{chapters/02-LayAdd/02-04-06}
	\input{chapters/02-LayAdd/02-04-07}
\input{chapters/02-LayAdd/02-05}
\input{chapters/02-LayAdd/02-06}
\input{chapters/02-LayAdd/02-07}