\chapter{Banana terms}\label{ch:BanTer}
Many different words
and etyma for ‘banana’ are found throughout
Linguistic Wallacea and beyond.
This appendix presents data from 391 languages/varieties.
This tally includes 329 Austronesian languages
and 62 Papuan languages.
192 of the Austronesian languages are within our target region.

Our primary focus is on reflexes of PMP *punti,
and on {\#}muku, the latter a pseudo-etymon with reflexes found
in several unrelated language families,
both Austronesian and Papuan.
The etymon of {\#}muku ‘banana’ is assumed to ultimately have
a non-Austronesian source (proto-Greater West Bomberai *maŋgʷVl
‘banana’; \citealp[478, 486]{UsherSchapper2022}),
and provides independent evidence that words can -- and have -- spread
widely through Austronesian languages in certain parts of the
target region without having to derive from a common Austronesian parent language.
(See discussion in \srf{02-03-01-07}
and \frf{15-11}.) We note that several
other etyma also have Austronesian-Papuan crossover (including
*punti), again indicating contact is a factor.

\frf{02-20} shows the distribution of reflexes of {\#}muku ‘banana’
in both Austronesian and Papuan languages.
\frf{02-20} also shows that reflexes of {\#}muku are found mostly
in the southern string of islands
and Bomberai, are not found in several Wallacean subgroups,
and are found only in certain branches of others.
Therefore contact has to be seen as an important part of the story of {\#}muku.
\frf{15-11} shows the distribution of several etyma for ‘banana’
which have more than one attestation in our target area and neighbouring areas.

This appendix provides additional data
and refinement to the seminal work of \citet{DonohueDenham2009}
for this and surrounding regions.
The etyma given in this section are mostly those given by \citet{DonohueDenham2009},
with several modifications and additions.
The different etyma identified by \citeauthor{DonohueDenham2009}
which occur east of the Blust line are summarised
in \trf{B-01}.\footnote{
		There are 391 languages in our sample,
		but 403 total terms for ‘banana’.
		This discrepancy is due to the fact that several languages
		have a reflex of more than one etyma.
		This includes languages (such as those in Sumba)
		with different reflexes having different semantics,
		or where one dialect of a language has one reflex
		but another dialect a different reflex.}
The other/\it{misc.} row in this table includes
etyma that only occur west of the Blust line (e.g. *sagiŋ),
those which have multiple attestations in \citet{DonohueDenham2009},
but less than three attestations in our sample,
as well as etyma that have only one attestation in \citeauthor{DonohueDenham2009}.

\begin{table}[h]
	\centering\tcp{Banana etyma}{B-01}
	\st{6pt}\begin{tabular}{llll}\lsptoprule
Etymon		&	\tsc{An}	&	non-\tsc{An}	&	total	\\	\midrule
*punti		&	114	&	4		&	118	\\
{\#}muku	&	70	&	27	&	97	\\
*kalauR		&	65	&	4		&	69	\\
*telewa		&	16	&	0 	&	16	\\
*biak			&	13	&	1		&	14	\\
*b[ao]le	&	0 	&	7		&	7	\\
*rando		&	5		&	0 	&	5	\\
*bief			&	4		&	0 	&	4	\\
%*bui			&	2		&	--	&	2	\\
\it{other/misc.}	
					&	54	&	19	&	73	\\	\midrule
					&	341	&	62	&	403	\\
		\lspbottomrule
	\end{tabular}
\end{table}

Among these etyma, we can distinguish between two variants
of the {\#}muku etymon,
shown in \trf{B-02}.
Firstly, there are words which are inherited from Greater West Bomberai *maŋgʷVl (e.g.
Mbaham \it{muˈŋgʷo}, Klon \it{mogal}) or borrowed from one
of these inheritances (e.g. Uruangnirin \it{maŋgo},
Erokwanas and Bedoanas \it{maʔo}).\footnote{
		Erokwanas and Bedoanas
		have **g > \it{ʔ} (see \srf{11-05-01},
		\trf{11-30}), thus pointing to earlier **mago.}
These words
are found in \tsc{West Bomberai} languages
and some of their Austronesian neighbours,
as well as the \tsc{Alor-Pantar} Papuan languages.
Secondly, there are words which more directly reflect a form
like *mugu or *muku (e.g.
Lio \it{muku}, Ili{\Q}uun \it{muu},
Batuley \it{mug,} Makasae \it{muʔu}).
These words occur in Austronesian languages,
as well as Bunak
and the \tsc{East Timor} branches of \tsc{Timor-Alor-Pantar}.\footnote{
		See \citet[493]{UsherSchapper2022} for more discussion.
		They propose Proto-East Timor *mugu as a
		regular reflex of Proto-Greater West Bomberai *maŋgʷVl.
		However, some of the words in East Timor languages could be secondary
		borrowings from their Austronesian neighbours.}

\begin{table}[h]
	\centering\tcp{Variants of {\#}muku}{B-02}
	\st{6pt}\begin{tabular}{llll}\lsptoprule
&	\tsc{An}	&	non-\tsc{An}	&	total	\\	\midrule
*maŋgʷVl		&	3		&	22	&	25	\\
*muku/*mugu	&	56	&	5		&	61	\\
		\lspbottomrule
	\end{tabular}
\end{table}

\newpage
Similarly, there are three kinds of variants/reflexes of the *kalauR etymon.
These are shown in \trf{B-03}.
\citet{DonohueDenham2009} instead propose *kaloy which they connect
ultimately with PMP *qaRutay ‘plant in the banana family’ with
several intermediate, irregular sound changes.
However, we do not find the proposed links to *qaRutay entirely plausible.\footnote{
		\citet{DonohueDenham2009}
		propose *qaRutay > *kalutay > *kaluay > *kaloy > *kola > *loka.}
We also find that *kalauR (rather than *kaloy) better accounts
for the reflexes seen in the region.\footnote{
		Final *R accounts
		straightforwardly for all the forms in our region,
		while final *y does not.
		\tsc{Celebic} Munan languages (e.g.
		Muna \it{kalei} ‘banana’) attest regular *R > **y (see \srf{14-04-03-01}).
		Aru languages (e.g.
		Mariri \it{kalaur} ‘banana’) attest regular *R > \it{r} (see
		\srf{06-01-01}). Sumba languages (e.g.
		Kambera \it{kaluu} ‘banana’) attest *R > {\0} (see \srf{03-03}).
		The trisyllabic Sumba reflexes probably also have final vowel epenthesis (e.g.
		intermediate **kaloR-u, see \srf{03-03-01}) before loss of *R.}
Some languages
in our sample appear to directly reflect *kalauR.
Some languages clustered around Ambon (as well as Banda) have
*kula, with vowel metathesis and raising of the vowel.\footnote{
		A reviewer questioned the connection between the forms that
		we link together in \trf{B-03}.
		However, similar vowel and consonant metatheses that are required
		to derive *kula and *loka from *kalauR (via intermediate **kalo) are attested
		sporadically throughout the region.}
Other languages have *loka, with consonant metathesis.
Such languages occur in Sulawesi
and Halmahera.

\begin{table}[h]
	\centering\tcp{Variants of *kalauR}{B-03}
	\st{6pt}\begin{tabular}{llll}\lsptoprule
&	\tsc{An}	&	non-\tsc{An}	&	total	\\	\midrule
*kalauR	&	33	&	3		&	36	\\
*kula		&	7		&	0		&	7	\\
*loka		&	25	&	1		&	26	\\
\lspbottomrule
	\end{tabular}
\end{table}

The following sections list all the banana terms in our sample.
The sections follow the order of subgroups in this book,
going from west to east along the southern string of islands,
then west to east through central Maluku
and Bomberai, followed by \tsc{Celebic}
and other Austronesian languages further to the west
and north of Wallacea,
and then Papuan languages.
Within each subgroup the ordering is a compromise between node
and geography with nodes going from west to east and/or north to south.
Languages within a single node are grouped together,
and are similarly ordered from west to east and/or north to south.
All data in this appendix which is taken from \citet[312--332]{DonohueDenham2009}
is abbreviated to “DD”.

\input{chapters/B-BanTer/B-01}
\input{chapters/B-BanTer/B-02}
\input{chapters/B-BanTer/B-03}
\input{chapters/B-BanTer/B-04}
\input{chapters/B-BanTer/B-05}
\input{chapters/B-BanTer/B-06}
\input{chapters/B-BanTer/B-07}
\input{chapters/B-BanTer/B-08}
\input{chapters/B-BanTer/B-09}
\input{chapters/B-BanTer/B-10}
\scp{Sulawesi}{B-11}
	\input{chapters/B-BanTer/B-11-01}
	\input{chapters/B-BanTer/B-11-02}
	\input{chapters/B-BanTer/B-11-03}
	\input{chapters/B-BanTer/B-11-04}
	\input{chapters/B-BanTer/B-11-05}
	\input{chapters/B-BanTer/B-11-06}
\input{chapters/B-BanTer/B-12}
\scp{Papuan (non-Austronesian)}{B-13}
	\input{chapters/B-BanTer/B-13-01}
	\input{chapters/B-BanTer/B-13-02}
	\input{chapters/B-BanTer/B-13-03}