\chapter{Other subgrouping possibilities}\label{ch:OthSub}
In this chapter we expand on some issues
and evaluate a number of proposals that have been put forward
by others as having subgrouping value in the region,
as well as explore some additional possibilities.

In reviewing the evidence for potential higher level links,
we find little convincing evidence on the basis of the historical
phonology for linking the Wallacean subgroups discussed in the
previous chapters of this book together in any known intermediate
node below the level of PMP.
This includes lack of evidence from the historical phonology
in the region for traditionally accepted CMP and CEMP subgroups.
Some of the following issues relate to claims that have been
made about CMP, and others that have been made about CEMP.
Each is flagged according to the group to which the proposal relates.
While there may or may not be one or more intermediate nodes
between (some) Wallacean subgroups
and PMP, such putative nodes are yet to be identified.

In reviewing the study of subgrouping in the target region,
we track the claims
and counterarguments for some issues,
and note that while several claims have since been withdrawn
by their originators in the face of credible counterarguments,
nevertheless the earlier conclusions based on them have persisted in the literature.

In the previous chapters we have shown there is a complex chain
of subgroups in the region,
some of which share features in their historical phonologies with neighbouring subgroups.
Chapter \ref{ch:CM} discusses such chaining in some detail for subgroups
in the central Maluku region.
Features that are shared across some subgroups,
such as mergers *a/*-ay > **a,
and *n/*ñ > \it{n},
or a weakening of *p > \it{ɸ,
f, h}, {\0}, are not found in all subgroups,
often have exceptions, are also found commonly outside the target
region, and are so widespread as to not be given much weight for subgrouping.

We begin with a number of issues dealing with the historical
phonology (\srf{13-01}), expanding into phonotactic issues relating to the CV structures
of morphemes, words and phrases (\srf{13-02}), then look at issues in the morphosyntax (\srf{13-03}).
In \srf{13-04}, we take a bird’s-eye view to some additional
possibilities for combining various Wallacean subgroups based
primarily on the historical phonology.

\input{chapters/13-OthSub/13-01}
	\input{chapters/13-OthSub/13-01-01}
	\input{chapters/13-OthSub/13-01-02}
	\input{chapters/13-OthSub/13-01-03}
	\input{chapters/13-OthSub/13-01-04}
	\input{chapters/13-OthSub/13-01-05}
		\input{chapters/13-OthSub/13-01-05-01}
		\input{chapters/13-OthSub/13-01-05-02}
		\input{chapters/13-OthSub/13-01-05-03}
\input{chapters/13-OthSub/13-02}
	\input{chapters/13-OthSub/13-02-01}
		\input{chapters/13-OthSub/13-02-01-01}
		\input{chapters/13-OthSub/13-02-01-02}
	\input{chapters/13-OthSub/13-02-02}
	\input{chapters/13-OthSub/13-02-03}
\input{chapters/13-OthSub/13-03}
	\input{chapters/13-OthSub/13-03-01}
	\input{chapters/13-OthSub/13-03-02}
	\input{chapters/13-OthSub/13-03-03}
	\input{chapters/13-OthSub/13-03-04}
	\input{chapters/13-OthSub/13-03-05}
\input{chapters/13-OthSub/13-04}