\sscp{\tsc{East Bomberai}}{11-05-02}
The \tsc{East Bomberai} node of \tsc{Tanimbar-Bomberai} contains
\ili{Irarutu}, \ili{Kuri} (a.k.a \ili{Nabi}), \ili{Arguni}, \ili{Bedoanas} (a.k.a.
\ili{Yarik}), and \ili{Erokwanas} (a.k.a. \ili{Kambran}, \ili{Damberang-Goras}).
The \tsc{East Bomberai} languages have had extensive contact
with neighbouring Papuan languages,
in particular \ili{Iha} and \ili{Mbaham}.
This contact has resulted in borrowing in both directions
\citep{UsherSchapper2018,UsherSchapper2022}.
\citet{UsherSchapper2018} focus on borrowing from the Austronesian
languages into the Papuan languages,
but borrowings in the other direction also occur.
In particular, \citet{UsherSchapper2022} include the highly significant
Proto-Greater West Bomberai reconstructions *mandəl ‘bat’
and *maŋgʷVl ‘banana’ discussed in
\srf{14-04-04} and \srf{02-03-01-07} respectively.
Two examples which are more locally restricted are \ili{Iha} \it{hɛir},
\ili{Mbaham} \it{sey(i)r} > \ili{Arguni} \it{sair},
\ili{Bedoanas}, \ili{Erokwanas} \it{sai} all ‘fish’,
as well as \ili{Iha} \it{tɔqar},
\ili{Mbaham} \it{tɔːqar} both ‘bone’ > \ili{Arguni} \it{tór},
\ili{Bedoanas} \it{tóʔor}, \ili{Erokwanas} \it{toʔar} all ‘thorn,
bone’, and \ili{Irarutu} \it{tor} [torə] ‘bone’.\footnote{
		\ili{Arguni} \it{tór},
		\ili{Bedoanas} \it{tóʔor}, \ili{Erokwanas} \it{toʔar} all ‘thorn,
		bone’, and \ili{Irarutu} \it{tor} [torə] ‘bone’ might alternately be
		from PMP *tuqəlan ‘condylar bone; bone of fauna exclusive of
		fish’, though we would expect *q > {\0} (not *q > \it{ʔ}) in \ili{Bedoanas}
		and \ili{Erokwanas}.} 
The heavily Papuanised nature of these languages,
combined with scarcity of data until recent years,
has meant that the position of these languages has remained unclear.
\citet[272]{Blust1993} placed \ili{Arguni} with the \tsc{Bomberai Tip}
languages (and thus within his CMP) noting that it “[{\ldots}]
is the most divergent member of the group.” \citet{Blust1993}
left \ili{Irarutu} unclassified within CEMP,
and had no access to data from \ili{Bedoanas} or \ili{Erokwanas} to make any assessment.
\citet{Kamholz2014}, in his investigation of the SHWNG subgroup,
also considered the position of these languages focusing on \ili{Irarutu}.
He came to no firm conclusion but excluded them from his examination of SHWNG.

\largerpage
Evidence that these languages do not belong within SHWNG comes
from the reflexes of *j,
as exemplified in \trf{11-04} and \trf{11-22}.
SHWNG languages show a merger of *s/*j \citep[52f]{Kamholz2014}.
Unlike SHWNG languages, *j
and *s do not usually merge in \tsc{East Bomberai}.
The reflexes of *j are discussed in more detail in the following sections.
Evidence that these languages belong within STB comes from shared
merger of *z/*d > **d (\srf{11-01-01}).

\begin{table}\tcp{\tsc{East Bomberai} *l > \it{r}}{11-30}
%\begin{adjustbox}{width=\textwidth}
	\begin{threeparttable}\st{6pt}
	\begin{tabular}{llllll}\lsptoprule
*gloss	&	sky	&	head	&	ten	&	root	&	road, path\\
PMP	&	*\tbr{l}aŋit	&	*qu\tbr{l}uh	&	*saŋapu\tbr{l}uq	&	*da\tbr{l}ij\su{Δ}	&	*za\tbr{l}an\\	\midrule
\ili{Arguni}	&	{\hr}\tbr{r}ágit	&	{\hr}á\tbr{r}u-g	&	{\hr}sambú\tbr{r}e	&	{\hr}da\tbr{r}e(g)	&	{\hr}rá\tbr{r}in\\
\ili{Bedoanas}	&	{\hr}\tbr{r}aʔit	&	{\hr}i\tbr{r}i atu\su{†}	&	{\hr}sambú\tbr{r}a	&	{\hr}da\tbr{r}a, nda\tbr{r}ao	&	{\hr}rá\tbr{r}in\\
Erokw.	&	{\hr}\tbr{r}áʔit	&	{\hr}a\tbr{r}u puat\su{†}	&	{\hr}sambú\tbr{r}a	&	{\hr}dá\tbr{r}a	&	{\hr}ró\tbr{r}in\\
\ili{Irarutu}	&	{\hr}\tbr{r}agt	&	{\hr}\tbr{r}ü \su{‡}	&	\cg{(fradru)}	&	\cg{(war)}	&	{\hr}ra\tbr{r}n\\
\ili{Kuri}	&		&	\cg{(tə̆bon)}	&	\cg{(βradu)}	&	\cg{(war)}	&	{\hr}ra\tbr{r}n\\
		\lspbottomrule
	\end{tabular}
		\begin{tablenotes}\tnsize
			\item[†]	{\ili{Bedoanas} \it{atu}
								and \ili{Erokwanas} \it{puat} are probably from *batu ‘stone’.}
			\item[‡]	{\ili{Irarutu} \it{rü} is glossed ‘source’ \citep[222]{Jackson2014}.
								The vowel is a front rounded vowel.}
			\item[Δ]	{Reflexes of *dalij have *d = \it{d} against normal *d > \it{r}.
								However, compare also the reflexes of *z (\trf{11-03}) for which one two forms
								have *z > \it{r} and one has *z > \it{d}.
								\ili{Arguni} \it{dareg} is from \cite{Narfafan2011} and \it{dare}
								from \citet[215]{SmitsVoorhoeve1992a}.}
		\end{tablenotes}
	\end{threeparttable}
%\end{adjustbox}
\end{table}

The main defining sound change of \tsc{East Bomberai} is shared
*l > \it{r}.\footnote{
		The North\ili{east} \ili{Arguni} dialect of \ili{Irarutu}
		has subsequent **r > \it{l} \citep[8]{Jackson2014}.}
This change also occurs in neighbouring \ili{Kowiai} (\srf{11-03-02}).
Examples are given in \trf{11-25},
with an additional four examples given later in \trf{11-30}.
To these we can add a number of words for which only the \ili{Arguni}
reflex is currently known: *kapa\tbr{l} > \it{maʔápe\tbr{r}}
‘thick’ (with irregular *p = \it{p}),
*qa\tbr{l}ipan > \it{a\tbr{r}ifen} ‘centipede’,
and *tə\tbr{l}ən > \it{ute\tbr{r}en} ‘swallow’.

\begin{table}\tcp{\tsc{East Bomberai} *ŋ > \it{g}}{11-31}
\begin{adjustbox}{width=\textwidth}
	\begin{threeparttable}\st{2pt}
	\begin{tabular}{lllllll}\lsptoprule
*gloss	&	name	&	cry	&	ear	&	finger, toe	&	branch	&	sky\\
PMP	&	*\tbr{ŋ}ajan	&	*ta\tbr{ŋ}is	&	*ta\tbr{ŋ}ila/*tali\tbr{ŋ}a	&	*ta\tbr{ŋ}an	&	*sa\tbr{ŋ}a	&	*la\tbr{ŋ}it\\	\midrule
\ili{Yamdena}	&	{\hr}\tbb{ŋ}ar-e	&	{\hr}-tasi\tbb{ŋ}	&	{\hr}tli\tbb{ŋ}a-\tbb{ŋ}w	&	{\hr}ta\tbb{ŋ}a-ny	&	{\hr}sa\tbb{ŋ}a-ny	&	{\hr}la\tbb{ŋ}it\\	\midrule
\ili{Fordata}	&	{\hr}\tbb{n}ara-\tbb{n}	&	{\hr}\tbb{n}-ta\tbb{n}it	&	\cg{(aru-n)}	&		&	{\hr}a\tbb{n}a-\tbb{n}	&	{\hr}la\tbb{n}it\\
\ili{Kei}	&	\cg{(mema-n)}	&	{\hr}tna\tbb{n}it	&	\cg{(aru-n)}	&		&	{\hr}ha\tbb{ŋ}a	&	{\hr}la\tbb{n}it\\
\ili{Onin}	&	{\hr}\tbr{g}ara-r	&	\cg{(a-gaga)}	&	{\hr}tani\tbr{g}a-n	&	{\hr}nemi-n ta\tbr{k}an	&		&	\cg{(laŋit)}\su{†}\\
\ili{Sekar}	&	{\hr}\tbr{g}ara-r	&	{\hr}a-ta\tbr{g}is	&	{\hr}tani\tbr{g}a-n	&	{\hr}nima-n ta\tbr{g}an	&		&	\cg{(laŋit)}\su{†}\\
Uruang.	&	{\hr}(\tbr{ŋ})\tbr{g}aran 	&	\cg{(a-rekar)}	&	{\hr}tani\tbr{ŋ}a-n	&	{\hr}ta\tbr{ŋ}karek	&	{\hr}ra\tbr{ŋg}as	&	{\hr}na\tbr{g}it\su{†}\\	\midrule
\ili{Arguni}	&	{\hr}\tbr{g}ádin	&	{\hr}t\<y\>á\tbr{g}īs	&	{\hr}ti\tbr{g}yánáne	&	{\hr}tá\tbr{g}e	&	{\hr}sá\tbr{g}e	&	{\hr}rá\tbr{g}it\\
\ili{Bedoanas}	&	{\hr}\tbr{g}adiŋ	&	{\hr}t\<i\>á\tbr{ʔ}is	&		&	\cg{(baʔya)}	&		&	{\hr}ra\tbr{ʔ}it\\
Erokw.	&	{\hr}\tbr{g}adin	&	{\hr}t\<i\>á\tbr{ʔ}is	&	{\hr}ti\tbr{ʔ}inyanya	&	\cg{(rumwa)}	&		&	{\hr}rá\tbr{ʔ}it\\
\ili{Irarutu}	&	\cg{(no)}	&	{\hr}n-ta\tbr{g}	&	{\hr}t\tbr{g}ra	&	\cg{(fra ntü)}	&	\cg{(ran)}	&	{\hr}ra\tbr{g}t\\
\ili{Kuri}	&	\cg{(inɔi)}	&	{\hr}itə-ta\tbr{g}	&	{\hr}tə̆\tbr{g}ra	&		&	{\hr}sa\tbr{g}a	&	\\
		\lspbottomrule
	\end{tabular}
		\begin{tablenotes}\tnsize
			\item[†] {The \ili{Onin} and \ili{Sekar} words for ‘sky’ are \ili{Malay} borrowings.
								\ili{Uruangnirin} \it{nagit} ‘sky’ is from \citet[22]{SmitsVoorhoeve1992b}.
								\citet{Visser2023} has \it{laŋit}.}
		\end{tablenotes}
	\end{threeparttable}
\end{adjustbox}
\end{table}

Another consonant change shared between the \tsc{East Bomberai}
languages is *ŋ > \it{g} with subsequent *g > \it{ʔ} in \ili{Bedoanas} and \ili{Erokwanas}.
Examples are given in \trf{11-31}.
This change is not an exclusively shared innovation.
It also occurs in \tsc{Bomberai Tip} languages.
\ili{Onin} and \ili{Sekar} have *ŋ > \it{g}
and \ili{Uruangnirin} has *ŋ > \it{ŋg, g, ŋ}.
\ili{Kowiai} also has *g > \it{ŋg, g} (\srf{11-03-02}),
\ili{Teor} has *ŋ > \it{g} (\srf{11-04}),
and \tsc{East Seram} has similar medial *ŋ
> \it{k} (\srf{11-02}).

\tsc{East Bomberai} languages usually share word-initial *b >
**p, with subsequent **p > {\0} in \ili{Bedoanas}.
\ili{Irarutu} has subsequent **b > \it{f} thus indicating a merger of PMP *b/*p.
\ili{Kuri} also appears to merge *b/*p with a variety of realisations
(probably allophonic) including: [b],
[β], [f], and [v].
Whether the merger in \ili{Kuri} can be reconciled with *b > **p posited
for other languages is unclear.
Examples of PMP *b in \tsc{East Bomberai} are given in \trf{11-32}.
\ili{Onin} and \ili{Uruangnirin} of \tsc{Bomberai Tip} also have *b > \it{p.}

\begin{table}
	\centering\tcp{\tsc{East Bomberai} *b}{11-32}
	\st{6pt}\begin{tabular}{llllll}\lsptoprule
*gloss	&	moon	&	body hair, feather	&	new	&	stone	&	bow\\
PMP	&	*\tbr{b}ulan	&	*\tbr{b}ulu	&	*\tbr{b}aqəRu	&	*\tbr{b}atu	&	*\tbr{b}usuR\\	\midrule
\ili{Arguni}	&	{\hr}\tbr{p}úrin	&	{\hr}pu-\tbr{p}ure	&	{\hr}pu-\tbr{p}uaru	&	{\hr}\tbr{p}uát	&	{\hr}u\tbr{p}úsir\\
\ili{Bedoanas}	&	{\hr}úrin	&	{\hr}ur-ura	&	{\hr}ur-uri	&	{\hr}uat	&	\\
Erokw.	&	{\hr}\tbr{p}úrin	&	{\hr}pu-\tbr{p}ura	&	{\hr}\tbr{p}uari	&	{\hr}\tbr{p}uat	&	\\
\ili{Irarutu}	&	\cg{(seba)}	&	{\hr}\tbr{f}ru	&	\cg{(bunat)}	&	\cg{(kami)}	&	\cg{(fae)}\\
\ili{Kuri}	&	\cg{(sarabɛ)}	&	{\hr}ˈ\tbr{β}aru, \tbr{b}aru	&	\cg{(aˈbɔb)}	&	\cg{(kɔbaːb)}	&	\cg{(uˈagubɔ)}\\
		\lspbottomrule
	\end{tabular}
\end{table}

In addition to such examples of *b > **p,
there are also at l\ili{east} two examples of *b > \it{(m)b} in \ili{Arguni},
\ili{Bedoanas}, and \ili{Erokwanas} in our data: *\tbr{b}abuy/*\tbr{b}əRək
‘pig’> \ili{Arguni}, \ili{Bedoanas} \it{\tbr{mb}}o, \ili{Erokwanas} \it{\tbr{b}o},
and *\tbr{b}uaq ‘fruit’ > \ili{Arguni} \it{\tbr{mb}ú},
\ili{Bedoanas}, \ili{Erokwanas} \it{\tbr{mb}u} (\ili{Irarutu} \it{fu} ‘fruit’).
The reflexes of word medial *b are unclear due to scarcity of examples.
There is one example of *b > \it{p}; *Ri\tbr{b}u ‘thousand’ >
\ili{Arguni} \it{ri\tbr{p}isye}, \ili{Bedoanas} \it{ri\tbr{p}isa},
and one example of *b > \it{f}; *tə\tbr{b}uh
> \ili{Arguni} \it{to\tbr{f}} ‘sugarcane’.

\ili{As} in many languages of the region reflexes of *ma-p in \tsc{East
Bomberai} are different to those of *b and *p.
Only two items have been identified: *ma-putiq ‘white’ > \ili{Irarutu}
\it{\tbr{bf}ut} [\tbr{mbəf}ut], \ili{Kuri} \it{\tbr{maβ}utə̆}, as well as *ma-panas ‘hot,
warm’ > \ili{Arguni} \it{mba{\tl}\tbr{mb}anes},
\ili{Bedoanas}, \ili{Erokwanas} \it{ba{\tl}\tbr{b}anas,} and perhaps \ili{Irarutu}
\it{\tbr{w}in}, \ili{Kuri} \it{a\tbr{β}en}.

The reflexes of final vowels also provide evidence for \tsc{East Bomberai}.
Most final syllables are deleted in these languages.
However, they also have epenthesis of final vowels after word
final consonants to various extents.
Thus, loss of final vowels might be best viewed as reduction
rather than complete deletion.
Final syllable deletion (or reduction)
and epenthesis of final vowels are more common in \tsc{\ili{Irarutu}-Kuri}
than in other \tsc{East Bomberai} languages.
Final vowel epenthesis in \ili{Irarutu} is described by \citet[653]{Voorhoeve1995a}
as “[{\ldots}] a weak vocalic release.
Usually this is a mid-central vowel [ə],
but it can also be a weak echo-vowel”.
Examples of deletion of final vowels in \tsc{East Bomberai} are given in \trf{11-33}.

\begin{table}\tcp{\tsc{East Bomberai} *V > {\0} /{\gap}{\#}}{11-33}
%\begin{adjustbox}{width=\textwidth}
	\st{2.5pt}\begin{tabular}{llllllll}\lsptoprule
*gloss	&	male	&	die	&	banana	&	faeces	&	sugarcane	&	louse	&	br\ili{east}\\
PMP	&	*maRuqan\tbr{ay}	&	*mat\tbr{ay}	&	*punt\tbr{i}	&	*taq\tbr{i}	&	*təb\tbr{u}h	&	*kut\tbr{u}	&	*sus\tbr{u}\\	\midrule
\ili{Arguni}	&	{\hr}máran	&	{\hr}ga-mat	&	{\hr}fúd	&	{\hr}tā	&	{\hr}tof	&	{\hr}út	&	{\hr}sus\\
\ili{Bedoanas}	&	{\hr}máran	&	{\hr}mú-mat	&	\cg{(maʔo)}	&		&		&	{\hr}ota	&	\cg{(kókon)}\\
Erokw.	&	{\hr}máran	&	{\hr}í-m\<y\>at	&	\cg{(maʔo)}	&	{\hr}ta	&		&	{\hr}óta	&	\cg{(kákon)}\\
\ili{Irarutu}	&	{\hr}mran	&	{\hr}n-mat	&	{\hr}fud	&	{\hr}ta	&	{\hr}tof	&	{\hr}ut	&	{\hr}sus\\
	&	[məranə]	&	[nəmatə]	&	[fudə]	&	[ta]	&	[tofə]	&	[utə]	&	[susə]\\
\ili{Kuri}	&	{\hr}bran	&	{\hr}inə-mat	&	\cg{(wɛˈrɔm)}	&		&	\cg{(wɛtit)}	&	\cg{(təni)}	&	{\hr}susɛ\\
		\lspbottomrule
	\end{tabular}
%\end{adjustbox}
\end{table}

Examples of epenthesis of final vowels in \tsc{East Bomberai}
languages (apart from \ili{Irarutu}
and \ili{Kuri}) include: *Ratus ‘hundred’ > \ili{Arguni} \it{rátis\tbr{e}},
\ili{Bedoanas} \it{rátis\tbr{a}}, \ili{Erokwanas} \it{ratis\tbr{a}},
*kutu ‘louse’ > **ut > \ili{Arguni} \it{út},
\ili{Bedoanas}, \ili{Erokwanas} \it{ot\tbr{a}}, *sa-ŋa-puluq ‘ten’ > **sambur
> \ili{Arguni} \it{sambúr\tbr{e}}, \ili{Bedoanas}, \ili{Erokwanas} \it{sambúr\tbr{a}},
and *bulu ‘body hair,
feather’ > **pur > \ili{Arguni},
\it{pu-pur\tbr{e}} \ili{Bedoanas} \it{ur-ur\tbr{a}}, \ili{Erokwanas},
\it{pu-pur\tbr{a}}.
