\chapter{Banda}\label{ch:Ban}
The Banda language is no longer spoken in the Banda Islands located
about 140 km south of Seram,
but is still spoken in two separated remnant refugee communities
of Banda-Elat and Banda-Eli almost 400 km to the southeast of Banda on the
island of Kei Besar (see \frf{11-02}).
This move happened after the population of Banda was decimated
by the Dutch in 1621,
and survivors fled their homeland \citep{CollinsKaartinen1998,Collins2003b}.
The eleven volcanic Banda Islands in the Banda Sea are isolated
enough to be the botanical source of nutmeg
and mace (\it{Myristica frangrans}),
and such geographical isolation may also have contributed to
the apparently unique linguistic development of the language.
Evidence of the presence of speakers of Austronesian languages
in the Banda Islands dates to around 3,500 BP \citep{Spriggs2011,Lape2018}.

Out in the middle of the Banda Sea,
the Banda Islands have \tsc{Ambon-Seram} languages to the north
and west (\crf{ch:AS}); \tsc{Seram-Tanimbar-Bomberai} languages
to the north and east (\crf{ch:STB}); \tsc{Aru} languages to the southeast
(\crf{ch:Aru}); and \tsc{Timor-Babar} languages to the south and southwest (\crf{ch:TB}).

\citet[182--190]{Stresemann1927} recognised that while the Banda
language, which he called “Eli-Elat”,
shared some vocabulary with his Proto-Ambon subgroup (roughly
our \tsc{Ambon-Seram} plus parts of \tsc{Sula-Buru}; see \crf{ch:CM}),
it had a different phonological history,
preserving features of an earlier stage of linguistic development.
Unlike Proto-Ambon, Banda preserved historical final *k,
preserved final consonants on verbs that have been lost in Proto-Ambon,
fully preserved both parts of medial *NC clusters,
and had many other differences from the languages of Proto-Ambon.
From a modern perspective,
however, many of those differences could reflect secondary developments
from a common source.
More importantly \citet[185, 188]{Stresemann1927}
also noticed different results of a merger of PMP *R/*d/*z
> \it{r} in Banda from Proto-Ambon *d/*z > **d,
and *R > \it{r}.
He also saw that Banda had a different reflex of *ns > \it{ʤ},
from *nt/*nd/*nz/*nj > \it{d,
nd}, and thus did not share the merger of *nt/*nd/*nz/*nj/*ns
> **ḏ purportedly found in Proto-Ambon.
However, perusal of his data shows that many of his forms are
derived from Malay loans,
and others are from morphophonemic processes on inflected verbs
(see \trf{12-01}), so we have not yet found unambiguous examples that support his
distinctions from the historical phonology.
\citeauthor{Stresemann1927} also felt Banda was distinct from his Proto-Goram
subgroup (roughly the \tsc{Eastern Islands} node of our \tsc{Seram-Tanimbar-Bomberai} subgroup).

\citet{Chlenov1976} placed Banda as its own branch of his “Ambon”
subgroup, based mostly on lexicostatistics.
That separated it from Geser
and Watubela, which he placed in his “South Moluccan” subgroup.

\citet{Collins1986} argued that Banda is classified in a branch
under the East Central Maluku branch of his Central Maluku subgroup,
and separate from his Nunusaku (in our terms,
not part of our \tsc{Ambon-Seram} subgroup).
He grouped Banda together with two languages (Geser
and Watubela) of our \tsc{Eastern Islands} node of \tsc{Seram-Tanimbar-Bomberai}
under his label “Proto-Banda”,
partly based on shared mergers of PMP *d/*D with *z/*Z (which,
now recognised as retentions,
are no longer seen as having subgrouping value),
along with several other proposed splits
and mergers, some of which have exceptions in our data.
The strongest basis for “Proto-Banda” was *R/*d/*z > \it{r} shared
between Banda and Geser, even though other parts of the historical phonology,
the vocabulary and morphology of those two languages appear to be quite different.
Collins himself (\citeyear{Collins1986}) documents many other ways in the
historical phonology that set Banda apart from the other \tsc{Eastern
Islands} languages with which he grouped Banda.

\citet[112]{LoskiLoski1989} treat Banda as an isolate,
finding it is equally (un)related to languages in four different
nodes of our \tsc{Seram-Tanimbar-Bomberai} subgroup,
based on lexical similarity
(see \crf{ch:CM}, footnote \ref{fn:BandaLexicalSimilarity},
\prf{fn:BandaLexicalSimilarity}).

In short, to date there has been no agreement among various investigators
regarding the classification of Banda.
We treat Banda as an isolate,
with this classification being tentative.
The innovations in the historical phonology we discuss in this
chapter are listed below.
While most of the patterns that are discussed in this chapter
individually do not provide a strong basis for separating Banda
from \tsc{Ambon-Seram} or \tsc{Seram-Tanimbar-Bomberai},
nevertheless put together as a bundle they present a fairly strong
case for treating Banda as an isolate, pending additional data.

\begin{itemize}
		\item	The merger of nasals *n/*ŋ/*ñ > \it{n} in Banda (\srf{12-01})
					is shared with \tsc{Ambon-Seram}.
					However, this is not exclusive,
					and the collective differences in the historical phonology,
					morphosyntax, and the lexicon lead us to conclude that this
					common merger very likely happened through parallel development,
					and is not an exclusively shared innovation defining a larger
					subgroup that includes Banda.
		\item	In the apicals,
					Banda merges *d/*z/*R > \it{r}
					and *j/*l > \it{l} in ways that contrast with the patterns of
					mergers of apicals in both \tsc{Ambon-Seram} and \tsc{Seram-Tanimbar-Bomberai} (\srf{12-02}).
					At a higher level,
					it may be possible to link \tsc{Ambon-Seram},
					\tsc{Seram-Tanimbar-Bomberai}, and Banda together on the basis
					of shared merger of *d/*z (\srf{07-03-03}, \srf{13-04}).
		\item	In the labials,
					\citet[133]{Collins1986} observed “the treatment of the labial
					stops distinguishes Banda from the descendants of Seran Laut”
					(his Seran Laut ≈ our \tsc{Eastern Islands} node of \tsc{Seram-Tanimbar-Bomberai}).
					This is discussed in \srf{12-03}.
		\item	\citet[133f]{Collins1986} observed that reflexes of initial
					(*ma-C) and other initial and medial *NC clusters also distinguishes
					Banda from “Seran Laut” (\srf{12-04}).
		\item	In the vowels,
					Banda has *-ay/*-aw > \it{a},
					but final *a shows alternations of \it{a {\tl} o} which are not
					attested with reflexes of *-ay/*-aw,
					indicating that we cannot posit a straightforward merger of *-ay/*-aw/*-a in Banda (\srf{12-05}).
					This pattern contrasts with both \tsc{Ambon-Seram} and \tsc{Seram-Tanimbar-Bomberai}.
					\tsc{Ambon-Seram} does not merge *-ay/*-aw > \it{a} (see \srf{10-02}),
					and neither subgroup shows the alternations of final *a found in Banda.
		\item	Also in the vowels,
					Banda has penultimate *ə > \it{e}
					and a split of ultimate *ə > \it{e,
					o} as conditioned by the penultimate vowel (\srf{12-05}).
					It is possible to posit a partial merger of *ə/*a > \it{o} in
					final syllables, though there is evidence that this could not have gone through intermediate **a.
					This behaviour of ultimate *ə distinguishes Banda from \tsc{Seram-Tanimbar-Bomberai}
					in which ultimate *ə > \it{a}.
		\item	Banda preserves PMP *k = \it{k} in all word positions in all identified reflexes.
					This distinguishes it from \tsc{Ambon-Seram} languages which
					generally show a weakening or loss of *k > \it{ʔ},
					{\0}, except in the morphophonemic processes of verb inflections in some languages.
		\item	The \tsc{Eastern Islands} node of \tsc{Seram-Tanimbar-Bomberai}
					preserves PMP *q in all positions (see \srf{11-03-01}).
					Banda loses PMP *q > {\0} in all positions.
					If they shared a common parent language,
					this loss could be from a secondary development for Banda.
\end{itemize}

In the morphosyntax,
like \tsc{Ambon-Seram} languages, Banda has multiple inflectional
paradigms in its verbs,
as exemplified in \trf{12-01}.
Unlike \tsc{Ambon-Seram} languages, Banda is not known to make the human/non-human distinction.
Data in \trf{12-01} are adapted from \citet[528f]{CollinsKaartinen1998}.

\begin{table}
	\centering\tcp{Banda subject indexing and morphophonemic alternations}{12-01}
	\st{4pt}\begin{tabular}{llllllll}\lsptoprule
paradigm	&	Pro	&	V-verb	&	f-verb	&	t-verb	&	k-verb	&	n-verb	&	s-verb\\
gloss	&		&	take	&	pay	&	plant	&	bite	&	angle (fish)	&	bathe\\
PMP	&		&	*alaq	&	*bayaD	&	*tanəm	&	*kitkit	&		&	\\	\midrule
\tsc{1sg}	&	ak	&	k-ala	&	\tbf{k}iar	&	\tbb{t}ano	&	\tbb{k}iki	&	\tbb{n}afu	&	\tbf{ʧ}eli\\
\tsc{2sg}	&	ka	&	m-ala	&	\tbr{mb}iar	&	\tbr{nd}ano	&	\tbr{m(b)}iki	&	\tbb{n}afu	&	\tbr{ɲʤ}eli\\
\tsc{3sg}	&	i	&	n-ala	&	\tbr{mb}iar	&	\tbr{nd}ano	&	\tbf{ŋ}iki	&	\tbb{n}afu	&	\tbr{ɲʤ}eli\\
\tsc{1pl.incl}	&	kito	&	t-ala	&	ti-\tbb{f}iar	&	ta-\tbb{t}ano	&	ti-\tbb{k}iki	&	ta-\tbb{n}afu	&	te-\tbb{s}eli\\
\tsc{1pl.excl}	&	kam	&	kam-ala	&	ki-\tbb{f}iar	&	ka-\tbb{t}ano	&	ki-\tbb{k}iki	&	ka-\tbb{n}afu	&	ke-\tbb{s}eli\\
\tsc{2pl}	&	kem	&	m-ala	&	\tbr{mb}iar	&	\tbr{nd}ano	&	\tbr{m(b)}iki	&	\tbb{n}afu	&	\tbr{ɲʤ}eli\\
\tsc{3pl}	&	si	&	r-ala	&	ri-\tbb{f}iar	&	ra-\tbb{t}ano	&	ri-\tbb{k}iki	&	ra-\tbb{n}afu	&	re-\tbb{s}eli\\
		\lspbottomrule
	\end{tabular}
\end{table}

Also in the morphosyntax,
Banda has two noun class distinctions of alienable
and inalienable nouns, reflected in two different possessive
constructions \citep[535f]{CollinsKaartinen1998}.
Such distinctions are also found in \tsc{Ambon-Seram}
and \tsc{Seram-Tanimbar-Bomberai} languages, but not \tsc{Sula-Buru} (see \srf{02-03-01-01}).

In the lexicon, Banda \it{meliri} ‘stand’ is from PMP *diRi,
like most (but not all) \tsc{Seram-Tanimbar-Bomberai} languages (see \trf{11-02}).
This contrasts with \tsc{Ambon-Seram} languages that take their
word for ‘stand’ from PMP *kədəŋ (see \trf{07-06}).
Since both are retentions,
this has no real subgrouping value for Banda.
Note that Banda *diRi > \it{meliri} ‘stand’ has irregular *d
> \it{l} in Banda.
While we might want to posit regressive dissimilation,
this does not occur in other cases of *dVRV; *daRaq
> \it{raro-} ‘blood’, *daRəq > \it{raro} ‘clay’,
*duRi > \it{riri} ‘bone’.

Like \tsc{Ambon-Seram} languages, Banda \it{utuɲa} ‘hundred’
is derived from PCM **utum,
whereas \tsc{Seram-Tanimbar-Bomberai} languages derive ‘hundred’
from PMP *Ratus (see \trf{07-06}).
Like \tsc{Sula-Buru} and \tsc{Ambon-Seram} languages, Banda \it{lea}
‘sun’ reflects a lexically specific loss of PMP *j in *qaləjaw ‘sun, day’.
PMP *j is mostly preserved in \tsc{Seram-Tanimbar-Bomberai} languages
in this etymon (see \trf{07-05}).
Almost uniquely in the region,
Banda gets its word \it{tilu-} ‘ear’ from PMP *tilu/*tuli ‘earwax’,
rather than from *taliŋa or *taŋila as found
in neighbouring subgroups and languages.

\input{chapters/12-Ban/12-01}
\input{chapters/12-Ban/12-02}
\input{chapters/12-Ban/12-03}
\input{chapters/12-Ban/12-04}
\input{chapters/12-Ban/12-05}
