\sscp{Bandicoots \it{(Peramelidae)}}{C-06-02}
Terms for bandicoots (\it{Peramelidae}) in \tsc{Seram-Tanimbar-Bomberai}
languages are only confidently known for four languages, given below.
Uruangnirin \it{yawat kawer} ‘bandicoot’
and Onin \it{yawar teroba} ‘bandicoot’ are literally ‘earth rat/mouse’,
in the same way the local Malay term for bandicoot,
\it{tikus tanah}, is ‘earth rat/mouse’ (\it{yawat}
and \it{yawar} = ‘rat,
mouse’, \it{kawer} and \it{roba} = ‘soil, earth’).\footnote{
		The Uruangnirin
		and Onin terms may be calques on the local Malay \it{tikus tanah}
		‘bandicoot’ = ‘earth rat/mouse’.
		However, the source of this local Malay term may itself be
		a calque on an indigenous language of Papua.
		The semantics indicate that
		parallel innovation is also likely.}

Apart from terms explicitly glossed ‘bandicoot’ \citet[237]{SmitsVoorhoeve1992a}
have an entry headed ‘marsupial’.
Which Malay terms were used to elicit the responses included here are unknown.
However, the Onin, Sekar,
and Arguni words are similar enough to Uruangnirin \it{yawat
kawer} and Onin \it{yawar teroba} to raise the suspicion
that they may also designate ‘bandicoot’.\footnote{
		\citet[237]{SmitsVoorhoeve1992a} also give Kowiai \it{indor}
		and Irarutu \it{yemoga}, \it{ʤemuge} under their entry for ‘marsupial’,
		both of which are independently verified to mean ‘cuscus’ (see
		\srf{C-06-01}).}
\\

\noindent{\wg\st{2pt}\begin{tabularx}{\linewidth}
{llll>{\raggedright\arraybackslash}X}
\lsptoprule
%Language	&	ISO	&	Term	&	Gloss	&	Source	\\	\midrule
%\endfirsthead
Language	&	ISO	&	Term	&	Gloss	&	Source	\\	\midrule
%\endhead
Kei Kecil	&	kei	&	halee	&	mouselike 	&	\cite{Travis2019}	\\	\hhline{~}
					&			&				&	marsupial\footnotemark	&	\hp{\,}	\\
Kei Besar	&	kei	&	‹helé›&	bandicoot	&	\cite[196]{Ellen1993}	\\
Uruangnirin	&	urn	&	yawat kawer;	&	bandicoot;	&	\cite{Visser2023} (lit. `earth rat')	\\	\hhline{~}
	&		&	yawat	&	mouse, rat	&	\hp{\,}	\\
Onin-1	&	oni	&	yawat	&	marsupial, mouse	&	SV \citeyear[237, 243]{SmitsVoorhoeve1992a}	\\
Onin-2	&	oni	&	yawar teroba	&	bandicoot	&	\cite{Donohue2010} (lit. `earth rat')	\\
Sekar	&	skz	&	kawat	&	marsupial	&	SV \citeyear[237]{SmitsVoorhoeve1992a}	\\
Arguni	&	agf	&	rawake	&	marsupial	&	SV \citeyear[237]{SmitsVoorhoeve1992a} (C. met.?)	\\
Erokwanas	&	erw	&	adobat	&	rat, marsupial	&	\cite{Donohue2010}	\\
\lspbottomrule
\end{tabularx}}
\footnotetext{The Indonesian gloss of Kei \it{halee}
is \it{tikus berpundi} `pouched mouse/rat'.
It is given with the following
encyclopaedic information: ``Its snout is placed on a young
girl's breasts to make them fashionably small.''}

%\noindent{\wg\st{2pt}\begin{xltabular}{\linewidth}
%{lllll>{\raggedright\arraybackslash}X}
%\lsptoprule
%%Language	&	ISO	&	Term	&	Gloss	&	Etymon	&	Source	\\	\midrule
%%\endfirsthead
%Language	&	ISO	&	Term	&	Gloss	&	Etymon	&	Source	\\	\midrule
%\endhead
%
%\lspbottomrule
%\end{xltabular}}