\newpage
\scp{\tsc{Timor-Babar}}{C-01}

\noindent{\wg\st{2pt}\begin{xltabular}{\linewidth}
{lllll>{\raggedright\arraybackslash}X}
\lsptoprule
%Language	&	ISO	&	Term	&	Gloss	&	Etymon	&	Source	\\	\midrule
%\endfirsthead
Language	&	ISO	&	Term	&	Gloss	&	Etymon	&	Source	\\	\midrule	\hhline{~}
\endhead
\ili{Funai} \ili{Helong}	&	heg	&	mauka,	&	cuscus	&	(**mantəR),	&	\hp{\,}	\\	\hhline{~}
	&		&	muti	&		&	\it{misc.}	&	\hp{\,}	\\	\hhline{~}
	&		&				&		&							&	\multirow{-3}{=}{\raggedright \cite[19]{Edwards2018c} (\it{mauka} borrowed from \ili{Meto})}	\\
Roi{\Q}is \ili{Amarasi}	&	(aaz)	&	urum	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes	\\
\ili{Kotos} \ili{Amarasi}	&	aaz	&	ukum	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes	\\
\ili{Amanuban}	&	(aoz)	&	ukum	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes	\\
\ili{Amanatun}	&	(aoz)	&	ukum	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes	\\
\ili{Amanatun}, \ili{Lotas}	&	(aoz)	&	aukum	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes	\\
\ili{Amanatun}, Toianas	&	(aoz)	&	akum	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes	\\
Fatule{\Q}u	&	(aoz)	&	maukaʔ/	&	cuscus	&	(**mantəR)	&	Edwards fieldnotes	\\	\hhline{~}
	&		&	maukuʔ	&		&		&	\hp{\,}	\\
Amfo{\Q}an	&	(aoz)	&	ukum	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes	\\
\ili{Timaus}	&	(aoz)	&	ʔmaukuʔ	&	cuscus	&	(**mantəR)	&	Edwards fieldnotes	\\
\ili{Kopas}	&	(aoz)	&	ʔmaukaʔ/	&	cuscus	&	(**mantəR)	&	Edwards fieldnotes	\\	\hhline{~}
	&		&	ʔmaukuʔ	&		&		&	\hp{\,}	\\
\ili{Ketun}	&	(aoz)	&	ʔmaukaʔ	&	cuscus	&	(**mantəR)	&	Edwards fieldnotes	\\
\ili{Ketun}, Nai{\Q}oni	&	(aoz)	&	ʔmakuʔ	&	cuscus	&	(**mantəR)	&	Edwards fieldnotes	\\
\ili{Molo}	&	(aoz)	&	‹mauku›	&	cuscus	&	(**mantəR)	&	\cite[316]{Middelkoop1972}	\\
\ili{Miomafo}	&	(aoz)	&	ʔmaukaʔ	&	cuscus	&	(**mantəR)	&	Edwards fieldnotes	\\
\ili{Miomafo}, Tuamese{\Q}	&	(aoz)	&	makuʔ	&	cuscus	&	(**mantəR)	&	Edwards fieldnotes	\\	
\ili{Insana}	&	(aoz)	&	ʔmaukaʔ	&	cuscus	&	(**mantəR)	&	Edwards fieldnotes	\\
\ili{Kusa-Manea}	&	(aoz)	&	akum	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes	\\
\ili{Fehan} \ili{Tetun}	&	tet	&	kmeda	&	cuscus	&	(**mantəR)	&	\cite[112]{Morris1984}	\\
East \ili{Tetun}	&	tet	&	meda	&	cuscus	&	(**mantəR)	&	\cite[140]{Morris1984}	\\
Waima{\Q}a	&	wmh	&	mide	&	cuscus	&	(**mantəR)	&	\cite{HimmelmannFreitas2006}	\\
Dadu{\Q}a	&	---	&	madah	&	cuscus	&	**mantəR	&	\cite[99]{Penn2006}	\\
\ili{Galolen}	&	gal	&	madar	&	cuscus	&	(**mantəR)	&	\cite{Schapper2011}	\\
Ili{\Q}uun	&	ilu	&	maʧa	&	cuscus	&	**mantəR	&	\cite[127]{DeJosselinDeJong1947}	\\
\ili{Tugun}	&	tzn	&	maʧar	&	cuscus	&	**mantəR	&	\cite{Hinton2023}	\\
\ili{Makuva}	&	lva	&	mate	&	cuscus	&	(**mantəR)	&	\cite[433]{vanEngelenhoven2009}	\\
\ili{Roma}	&	rmm	&	madar	&	cuscus	&	**mantəR	&	\cite[28]{Steven1991}	\\
\ili{Kisar}	&	kje	&	masla	&	cuscus	&	(**mantəR)	&	\cite{ChristensenChristensen-Grimes1990}	\\
\ili{Leti}	&	lti	&	mada	&	cuscus	&	**mantəR	&	\cite[423]{vanEngelenhoven2004}	\\
\ili{Moa}	&	lex	&	mada	&	cuscus	&	**mantəR	&	\cite[145]{Jonker1932}	\\
\ili{Wetan}	&	lex	&	mada	&	cuscus	&	**mantəR	&	Edwards fieldnotes	\\
East \ili{Damer}, \ili{Ilih}	&	dmr	&	madar	&	cuscus	&	**mantəR	&	Edwards fieldnotes	\\
East \ili{Damer}	&	dmr	&	madar	&	cuscus	&	**mantəR	&	\cite[241]{Blust1982}	\\
\ili{Teun}	&	tve	&	mata	&	cuscus	&	**mantəR	&	Edwards fieldnotes	(\ili{Wetan} loan?) \\
\ili{Nila}	&	nil	&	kuskusa	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes (Mly. loan)	\\
\ili{Serua}	&	srw	&	kuskusa	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes (Mly. loan)	\\
\ili{Dawera-Daweloor}	&	ddw	&	ku	&	cuscus	&	\it{misc.}	&	Edwards fieldnotes	\\
North \ili{Babar}	&	bcd	&	mad-a	&	cuscus	&	**mantəR	&	Edwards fieldnotes	\\
SE. \ili{Babar}	&	vbb	&	mɔt	&	cuscus	&	**mantəR	&	Edwards fieldnotes	\\
E. \ili{Marsela}	&	vme	&	mɔt-ɛ	&	cuscus	&	**mantəR	&	Edwards fieldnotes	\\
\ili{Emplawas}	&	emw	&	mutt-o	&	cuscus,	&	**mantəR	&	Edwards fieldnotes \\	\hhline{~}
	&		&		&	civet	&		&	 (semantic expansion)\\
Tela	&	tvm	&	mut-ɛ	&	cuscus	&	**mantəR	&	Edwards fieldnotes	\\
\ili{Seluwasan}	&	(sws)	&	mad-e	&	cuscus	&	**mantəR	&	Edwards fieldnotes	\\
\ili{Makatian}	&	(sws)	&	mad-e	&	cuscus	&	**mantəR	&	Edwards fieldnotes	\\
\ili{Selaru}	&	slu	&	mat	&	cuscus	&	**mantəR	&	\cite{CowardND}	\\
\lspbottomrule
\end{xltabular}}