\ssscp{\tsc{Eastern Celebic}}{C-09-04-03}

\noindent{\wg\st{2pt}\begin{xltabular}{\linewidth}
{lllll>{\raggedright\arraybackslash}X}
\lsptoprule
%Language	&	ISO	&	Term	&	Gloss	&	Etymon	&	Source	\\	\midrule
%\endfirsthead
Language	&	ISO	&	Term	&	Gloss	&	Etymon	&	Source	\\	\midrule
\endhead
\mc{6}{c}{\tsc{Taliabu}}		\\	\midrule	\hhline{~}
\ili{Soboyo}		&	tlv	&	geke	&	cuscus\footnotemark 	&	\it{misc.}	&	\cite[56]{Fortgens1921},	\\	\hhline{~}
					&			&				&					&							& Edwards fieldnotes	\\	\midrule	\hhline{~}
\mc{6}{c}{\tsc{\ili{Saluan}-Banggai}}	\\	\midrule	\hhline{~}
\ili{Bobongko}	&	bgb	&	kuseʔ		&	cuscus				&	**kusay			&	\cite[13]{Mead2018a}	\\
\ili{Banggai}		&	bgz	&	kuai;		&	cuscus;				&	(**kusay)		&	\cite[202]{SantosaSuyanto2018}	\\	\hhline{~}
					&			&	buŋes		&	dwarf cuscus 	&	\it{misc.}	&	\hp{\,}	\\
\end{xltabular}}
\footnotetext{
		\citet[56]{Fortgens1921} gives \ili{Soboyo} \it{geke} ‘klappermuis’ = ‘coconut mouse’.
		That the referent is cuscus is confirmed by Edwards’ consultants
		who gave orthographic ‹geke.e› for \ili{Ambon} \ili{Malay} ‘kusu’ = ‘cuscus’.
		Compare also \ili{Buru} \it{ik-geke} `go call cuscus at full moon'.}

\noindent{\wg\st{2pt}\begin{xltabular}{\linewidth}
{lllll>{\raggedright\arraybackslash}X}
\lsptoprule
%Language	&	ISO	&	Term	&	Gloss	&	Etymon	&	Source	\\	\midrule
%\endfirsthead
Language	&	ISO	&	Term	&	Gloss	&	Etymon	&	Source	\\	\midrule
\endhead
\ili{Balantak}	&	blz	&	kuse		&	cuscus				&	**kusay			&	\cite{BusenitzBradbury2016}	\\
\ili{Saluan}		&	loe	&	kuseʔ;	&	bear cuscus;	&	**kusay			&	\hp{\,}	\\	\hhline{~}
					&			&	bolotok	&	dwarf cuscus	&	\it{misc.}	&	\multirow{-2}{=}{\raggedright	\cite[117]{MeadPasanda2015}}	\\
\ili{Batui}			&	zbt	&	kuseʔ		&	cuscus				&	**kusay			&	\cite[117]{MeadPasanda2015}	\\	\midrule
\mc{6}{c}{\tsc{Southeastern Celebic}, \tsc{Bungki-Tolaki}}	\\	\midrule	\hhline{~}
\ili{Bungku}		&	bkz	&	kuhe		&	cuscus				&	**kusay			&	\cite[444]{Mead1998}	\\
\ili{Tolaki}		&	lbw	&	kuhe,		&	bear cuscus;	&	**kusay			&	\cite[444]{Mead1998};	\\	\hhline{~}
					&			&	βeŋu;		&	dwarf cuscus;	&	\it{misc.}	&	Edwards fieldnotes	\\	\hhline{~}
					&			&	puda		&	cuscus				&	\it{misc.}	&	Edwards fieldnotes	\\
\ili{Moronene}	&	mqn	&	kuhe		&	cuscus				&	**kusay			&	David Andersen (DM)	\\
\ili{Padoe}			&	pdo	&	kuhe		&	cuscus				&	**kusay			&	\cite[444]{Mead1998}	\\
\ili{Taloki}		&	tlk	&	kuhe;		&	bear cuscus;	&	**kusay			&	\cite{Mead2020d}	\\	\hhline{~}
					&			&	doalo		&	dwarf cuscus	&	\it{misc.}	&	\hp{\,}	\\
\ili{Kulisusu}	&	vkl	&	kuhe;		&	bear cuscus;	&	**kusay			&	\cite{Mead2019}	\\	\hhline{~}
					&			&	toŋali	&	dwarf cuscus	&	**taŋali		&	\hp{\,}	\\
\ili{Kodeoha}		&	vko	&	monea		&	cuscus				&	\it{misc.}	&	Edwards fieldnotes	\\
\ili{Mori Bawah}&	xmz	&	kuhe;		&	bear cuscus;	&	**kusay			&	\cite[51]{Esser1927}	\\	\hhline{~}
					&			&	toŋali	&	dwarf cuscus	&	**taŋali		&	\hp{\,}	\\
\ili{Wawonii}		&	wow	&	toŋali	&	dwarf cuscus	&	**taŋali		&	\cite[201]{FaridaDahruddin2008}	\\	\midrule	\hhline{~}
\mc{6}{c}{\tsc{Southeastern Celebic}, \tsc{\ili{Muna}-Buton}}		\\	\midrule	\hhline{~}
\ili{Busoa}			&	bup	&	mante		&	cuscus				&	**mantəR		&	\cite{vandenBerg2020}	\\
\ili{Cia-Cia}		&	cia	&	kuse,		&	bear cuscus,	&	**kusay, 		&	\hp{\,}	\\	\hhline{~}
					&			&	mante;	&	bear cuscus;	&	**mantəR		&	\hp{\,}	\\	\hhline{~}
					&			&	toŋali	&	dwarf cuscus	&	**taŋali		&	\multirow{-3}{=}{\raggedright \cite{Mead2015,Mead2017c,MeadTruong2021} (\it{kuse} in \ili{Wasambua})}	\\
\ili{Kaisabu}		&	cia	&	mante		&	cuscus				&	**mantəR		&	\cite{Mead2017a}	\\
\ili{Lasalimu}	&	llm	&	suʔe;		&	bear cuscus;	&	**kusay			&	\cite{Mead2018b} \it{(met.)}	\\	\hhline{~}
					&			&	toŋali	&	dwarf cuscus	&	**taŋali		&	\hp{\,}	\\
\ili{Muna}			&	mnb	&	kuhe;		&	bear cuscus;	&	**kusay			&	\hp{\,}	\\	\hhline{~}
					&			&	kampaʁi	&	dwarf cuscus	&	\it{misc.}	&	\multirow{-2}{=}{\raggedright \cite{SiduVanDenBerg1996}}	\\
S. \ili{Muna}		&	mnb	&	kuʔe		&	cuscus				&	**kusay			&	\cite{Atakasi2006}	\\
\ili{Pancana}		&	pnp	&	mante		&	bear/dwarf 		&	**mantəR		&	\cite{McDowell2010}	\\	\hhline{~}
					&			&					&	cuscus				&							&	\hp{\,}	\\
\ili{Kambowa}		&	ues	&	mante;	&	bear cuscus;	&	**mantəR		&	\cite{Mead2020b}	\\	\hhline{~}
					&			&	tonali	&	dwarf cuscus	&	**taŋali		&	\hp{\,}	\\
\ili{Kioko}			&	ues	&	mante;	&	bear cuscus;	&	**mantəR		&	\hp{\,}	\\	\hhline{~}
					&			&	xuloko	&	dwarf cuscus	&	\it{misc.}	&	\multirow{-2}{=}{\raggedright \cite[13]{Mead2020c} (\it{xuloko} possibly squirrel)}	\\
\ili{Kumbewaha}	&	xks	&	suʔe;		&	bear cuscus;	&	**kusay			&	\cite{Mead2017b} \it{(met.)}	\\	\hhline{~}
					&			&	toŋali	&	dwarf cuscus	&	**taŋali		&	\hp{\,}	\\
\ili{Wolowa}		&	xks	&	mante		&	cuscus				&	**mantəR		&	\cite{Mead2017c}	\\
\ili{Kaimbulawa}&	zka	&	mante		&	bear/dwarf 		&	**mantəR		&	\cite{McDowell2010}	\\	\hhline{~}
					&			&					&	cuscus				&							&	\hp{\,}	\\
\lspbottomrule
\end{xltabular}}
