\ssscp{Other/ambiguous}{C-10-03-04}

\noindent{\wg\st{2pt}\begin{tabularx}{\linewidth}
{lllll>{\raggedright\arraybackslash}X}
\lsptoprule
%Language	&	ISO	&	Form	&	gloss	&	Etymon	&	Source	\\	\midrule
%\endfirsthead
Language	&	ISO	&	Form	&	gloss	&	Etymon	&	Source	\\	\midrule
%\endhead
\ili{Isirawa}\footnotemark 	&	srl	&	iˈroro	&	marsupial	&	**kindoR	&	SV \citeyear[117]{SmitsVoorhoeve1998}	\\
\ili{Kalabra}	&	kzz	&	me	&	marsupial	&	**mantəR	&	SV \citeyear[118]{SmitsVoorhoeve1998} 	\\
\ili{Karon Dori}	&	kgw	&	teftiag	&	marsupial	&	\it{misc.}	&	SV \citeyear[118, 218, 228]{SmitsVoorhoeve1998}	\\
\ili{Kwerba}	&	kwe	&	ˈbeberiʧa	&	marsupial	&	\it{misc.}	&	SV \citeyear[117, 218, 228]{SmitsVoorhoeve1998}	\\
\ili{Mawes}\footnotemark[21]	&	mgk	&	kudur	&	marsupial	&	**kVndoR(a)	&	SV \citeyear[128]{SmitsVoorhoeve1994}	\\
\ili{Moraid}	&	msg	&	mänoho	&	marsupial	&	\it{misc.}	&	SV \citeyear[118, 219, 228]{SmitsVoorhoeve1998}	\\
\ili{Mpur}	&	akc	&	kerer	&	marsupial	&	\it{misc.}	&	SV \citeyear[118, 218, 228]{SmitsVoorhoeve1998}	\\
\ili{Orya}\footnotemark[21]	&	ury	&	sindor	&	marsupial	&	**kindoR	&	SV \citeyear[128]{SmitsVoorhoeve1994}	\\
\ili{Samarokena}	&	tmj	&	ˈsinifa	&	marsupial	&	\it{misc.}	&	SV \citeyear[117, 218, 228]{SmitsVoorhoeve1998}	\\
\ili{Seget}	&	sbg	&	bau	&	marsupial	&	\it{misc.}	&	SV \citeyear[118, 228]{SmitsVoorhoeve1998}	\\
\ili{Sougb}	&	mnx	&	meːsur	&	marsupial	&	**mantəR	&	SV \citeyear[118]{SmitsVoorhoeve1998}	\\
\ili{Tanahmerah}	&	tcm	&	kuehɛburi	&	marsupial	&	\it{misc.}	&	SV \citeyear[118]{SmitsVoorhoeve1994}	\\
\ili{Tunggare}	&	trt	&	begih, bɛgi	&	marsupial	&	\it{misc.}	&	SV \citeyear[117]{SmitsVoorhoeve1998}	\\
\ili{Maybrat}	&	ayz	&	wafu;	&	marsupial;	&	\it{misc.}	&	SV \citeyear[118]{SmitsVoorhoeve1998}	\\	\hhline{~}
				&			&	ʤifʤa;	&	marsupial;	&	\it{misc.}	&	\	\\	\hhline{~}
				&		&	difidiag	&	small marsupial	&	\it{misc.}	&		\\
\ili{Yawa}	&	yva	&	panaim	&	bandicoot	&	\it{misc.}	&	\cite[28]{JonesPaai1989}	\\
\lspbottomrule
\end{tabularx}}
\footnotetext{
\ili{Isirawa}, \ili{Mawes}, and \ili{Orya} are located further \ili{east} than \frf{14-03} extends.
}

%\noindent{\wg\st{2pt}\begin{xltabular}{\linewidth}
%{lllll>{\raggedright\arraybackslash}X}
%\lsptoprule
%
%\lspbottomrule
%\end{xltabular}}