\newpage
\ssscp{Tree kangaroos (\it{Dendrolagus})}{C-10-03-03}

\noindent{\wg\st{2pt}\begin{xltabular}{\linewidth}
{lllll>{\raggedright\arraybackslash}X}
\lsptoprule
%Language	&	ISO	&	Form	&	gloss	&	Etymon	&	Source	\\	\midrule
%\endfirsthead
Language	&	ISO	&	Form	&	gloss	&	Etymon	&	Source	\\	\midrule	\hhline{~}
\endhead
\ili{Duriankere}	&	dbn	&	kesyoro	&	tree wallaby	&	{\#}kisor(o)	&	SV \citeyear[228]{SmitsVoorhoeve1998}	\\
\ili{Inanwatan}	&	szp	&	isoro	&	tree wallaby	&	{\#}kisor(o)	&	SV \citeyear[219]{SmitsVoorhoeve1998}	\\
\ili{Kais}	&	kzm	&	kisoro	&	tree wallaby	&	{\#}kisor(o)	&	SV \citeyear[229]{SmitsVoorhoeve1998}	\\
\ili{Puragi}	&	pru	&	kamaʤi	&	tree wallaby	&	**mantəR	&	SV \citeyear[229]{SmitsVoorhoeve1998}	\\
\ili{Yawa}	&	yva	&	maisyaonda 	&	tree wallaby	&	\it{misc.}	&	\cite[28]{JonesPaai1989}	\\
\ili{Karon Dori}	&	kgw	&	katuwa	&	tree wallaby	&	\it{misc.}	&	SV \citeyear[228]{SmitsVoorhoeve1998}	\\
\ili{Konda}	&	knd	&	‹märet›	&	tree wallaby	&	**mantəR	&	SV \citeyear[228]{SmitsVoorhoeve1998}	\\
\ili{Kwerba}	&	kwe	&	koreya	&	tree wallaby	&	\it{misc.}	&	SV \citeyear[228]{SmitsVoorhoeve1998}	\\
\ili{Maybrat}	&	ayz	& katuo	&	tree kangaroo	&	\it{misc.}	&	\cite[300]{Dol2007}	\\
\ili{Moi}	&	mxn	&	kasyor	&	tree wallaby	&	{\#}kisor(o)	&	SV \citeyear[228]{SmitsVoorhoeve1998}	\\
\ili{Mor}	&	moq	&	ˈiror/iora	&	tree wallaby	&	{\#}kisor(o)	&	\cite{Donohue2010}/SV \citeyear[229]{SmitsVoorhoeve1998}	\\
\ili{Moraid}	&	msg	&	‹mähäbi›	&	tree wallaby	&	\it{misc.}	&	SV \citeyear[118, 228]{SmitsVoorhoeve1998}	\\
\ili{Mpur}	&	akc	&	a soma;	&	black wallaby;	&	\it{misc.}	&	SV \citeyear[228]{SmitsVoorhoeve1998}	\\
		&			&	kambu	&	grey tree wallaby	&	\it{misc.}	&	SV 	\\
\ili{Samarokena}	&	tmj	&	mawena	&	tree wallaby	&	\it{misc.}	&	SV \citeyear[117, 218, 228]{SmitsVoorhoeve1998}	\\
\ili{Seget}	&	sbg	&	suen	&	tree wallaby	&	\it{misc.}	&	SV \citeyear[118, 228]{SmitsVoorhoeve1998}	\\
\ili{Tanahmerah}	&	tcm	&	uaɛ	&	tree wallaby	&	\it{misc.}	&	SV \citeyear[229]{SmitsVoorhoeve1994}	\\
\ili{Tunggare}	&	trt	&	kura	&	tree wallaby	&	\it{misc.}	&	SV \citeyear[228]{SmitsVoorhoeve1998}	\\
\lspbottomrule
\end{xltabular}}