\chapter{Survey of marsupial terms}\label{ch:SurMar}
This appendix presents the full data available to us relating
to marsupial terms in Austronesian languages of Sulawesi
and Linguistic Wallacea, including parts of Indonesian west Papua.
It provides fuller data than the discussion in \crf{ch:Mar}.
Different variations of **kVndoR(a) ‘cuscus’ (\srf{14-03}),
and **mantəR ‘cuscus’ (\srf{14-04})
and formally close words are found in different Wallacean subgroups,
and also in Papuan languages.
502 terms for marsupials (and similar mammals)
are presented from 288 languages/varieties,
of which 231 are Austronesian,
and 57 are Papuan.

A special thanks goes to David Mead,
who did the bulk of the hard work in collecting marsupial terms
across the Sulawesi region,
and then kindly allowed us to use his unpublished work.
His data show several clear links between the forms found in
Sulawesi languages and those found to the \ili{east},
raising important methodological and subgrouping questions in the process.

This appendix mainly concerns the forms which are found.
We do not address the issue of loans in this appendix.
Contact across subgroups and families has clearly happened.
There are no known reflexes of any of these terms
found in \tsc{\ili{Bima}-Lembata} languages.
No words designating marsupials occur in that subgroup because
it lies outside the range of marsupials.

When a cognate set has multiple attestations in the region,
its presumed etymon is also given.
Etyma in brackets indicate irregular sound changes are required
perhaps pointing to loan distribution; e.g.
\ili{Buru} \it{tonal} ‘cuscus’ in \srf{C-04} is irregular from **taŋali thus
pointing to this term being a borrowing from a \tsc{\ili{Muna}-Buton}
language (see beginning of \srf{14-04-05} for more discussion).
When a term appears to have no cognates or only limited cognates
in adjacent languages this is indicated with \it{misc.} (miscellaneous).

In sources relating to Sulawesi languages,
DM = David Mead.
%Sources saying “\citep{GrimesC1990}” indicates that the named
%investigator filled in \citet{GrimesC1990} by hand
%and sent a copy to Grimes for use in comparative studies.
Much of our data comes from
\citet{SmitsVoorhoeve1992a,SmitsVoorhoeve1992b,SmitsVoorhoeve1994,SmitsVoorhoeve1998}.
Throughout this appendix these works are abbreviated as SV,
followed by the appropriate year and page number.

\input{chapters/C-SurMar/C-01}
\input{chapters/C-SurMar/C-02}
\input{chapters/C-SurMar/C-03}
	\input{chapters/C-SurMar/C-03-01}
	\input{chapters/C-SurMar/C-03-02}
	\input{chapters/C-SurMar/C-03-03}
	\input{chapters/C-SurMar/C-03-04}
	\input{chapters/C-SurMar/C-03-05}
\input{chapters/C-SurMar/C-04}
\input{chapters/C-SurMar/C-05}
\input{chapters/C-SurMar/C-06}
	\input{chapters/C-SurMar/C-06-01}
	\input{chapters/C-SurMar/C-06-02}
	\input{chapters/C-SurMar/C-06-03}
\input{chapters/C-SurMar/C-07}
\input{chapters/C-SurMar/C-08}
	\input{chapters/C-SurMar/C-08-01}
	\input{chapters/C-SurMar/C-08-02}
	\input{chapters/C-SurMar/C-08-03}
	\input{chapters/C-SurMar/C-08-04}
\input{chapters/C-SurMar/C-09}
	\input{chapters/C-SurMar/C-09-01}
	\input{chapters/C-SurMar/C-09-02}
	\input{chapters/C-SurMar/C-09-03}
	\input{chapters/C-SurMar/C-09-04}
		\input{chapters/C-SurMar/C-09-04-01}
		\input{chapters/C-SurMar/C-09-04-02}
		\input{chapters/C-SurMar/C-09-04-03}
	\input{chapters/C-SurMar/C-09-05}
		\input{chapters/C-SurMar/C-09-05-01}
\scp{Papuan (non-Austronesian)}{C-10}
	\input{chapters/C-SurMar/C-10-01}
		\input{chapters/C-SurMar/C-10-01-01}
		\input{chapters/C-SurMar/C-10-01-02}
	\input{chapters/C-SurMar/C-10-02}
	\input{chapters/C-SurMar/C-10-03}
		\input{chapters/C-SurMar/C-10-03-01}
		\input{chapters/C-SurMar/C-10-03-02}
		\input{chapters/C-SurMar/C-10-03-03}
		\input{chapters/C-SurMar/C-10-03-04}